---
title: "High dimensional immune profiling following aHSCT in MS - PART2"
author: "Dimitrios Kleftogiannis"
date: "2024-9-15"
output: html_document
---

### Utility

The data interpretation is performed in collaboration with **Dr. Sonia Gavasso and Jonas Haugsøen**. 

The utility of this code is to reproduce the part of the analysis with the cellular states as summarised in main Figures 4-5 of the manuscript.

### Contact

Comments and bug reports are welcome, please email: Dimitrios Kleftogiannis (dimitrios.kleftogiannis@uib.no)

We are also interested to know about how you have used our source code, including any improvements that you have implemented.
 
You are free to modify, extend or distribute our source code, as long as our copyright notice remains unchanged and included in its entirety. 

### License

This code is licensed under the MIT License.

Copyright 2024, University of Bergen (UiB) and Neuro-SysMed, Norway


### Loading packages 
First we load the packages required for the analysis.

```{r load packages, echo=FALSE, eval=FALSE, error=TRUE, warning=FALSE,cache=TRUE}
# set global chunk options
# load libraries
library(RColorBrewer)
library(gplots) 
library(ggplot2)
library(gdata)
library(gridExtra)
library(grid)
library(reshape2)
library(ggthemes)
library(ggpubr)
library(dplyr)
library(umap)
library(ComplexHeatmap)
library(ConsensusClusterPlus)
library(readxl)
library(igraph)
library(matrixStats)
library(ggforce)
library(ggraph)
library(ggridges)
library(vite)
library(MASS)
```

### Initialise working directory and load in-house functions
```{r initialise workspace,cache=TRUE,echo=FALSE}
setwd("/Home/siv32/dkl081/Desktop/RAMMS/code_repo")
```

### Specify the markers of interest in the antibody panel
```{r specify marker names,cache=TRUE,echo=FALSE}
#all markers
col.names <- c("CD16","CD4","CD14","CD19","CD3","CD235ab","CD11c","CD33",
               "CD133","CD123","CD162","CD185","CD45RA","CD278","CD194","CD161",
               "CD184","CD27","CD44","CD127","CD10","CD73","HLADR","CD146",
               "CD117","CD8a","CD34","CD105","CD49d","CD20","CD25","CD66b",
               "CD49f","CD45RO","CD90","CD45","CD195","CD38","CD196","CD135","CD56")
col.names


col.names.info <- c('type','type','type','type','type','state','type','type',
                    'state','type','state','state','type','state','state','state',
                    'state','state','state','state','state','state','type','state',
                    'state','type','type','state','state','type','state','type',
                    'state','type','state','type','state','type','state','state','type')

#define the marker_info
marker_info <- data.frame(Channel=NA,
                          marker_name=col.names,
                          marker_class=col.names.info)

```

### Load the cellular states found using unsupervised learning in a small training set of cells and predict the results for the remaining cells in the discovery cohort

```{r load results in the discovery cohort, echo=TRUE, cache=TRUE,eval=FALSE}

#discovery cohort
groupA <- 'BL'
groupD <- '100d'
groupE <- '6m'
groupF <- '12m'

#find the state markers
stateMarkers <- marker_info[marker_info$marker_class=='state',]
stateMarkerNames <- stateMarkers$marker_name

#we work with 17 cell types in this order
all_cellType_order <- c("CD4_Memory" ,"CD4_Naive" ,
                        "CD8_Memory","CD8_Naive",
                        "T_DN","T_other",
                        "B_Memory" ,"B_Naive" ,'B_other',
                        "NK_CD16_pos", "NK_CD16_neg", "NK_other",
                        "Monocytes_classical", "Monocytes_nonClassical", "DC","DC_plasmocytoid","Myelocytes_other")

colors_all <- c('slategray1','slategray3',
                'steelblue1','steelblue3',
                'darkslategrey','slateblue4',
                'firebrick1','firebrick4','indianred1',
                'orange','orange3','peachpuff',
                'palegreen3','palegreen1','lightgoldenrod3','lightgoldenrod1','greenyellow')

mySelectedCellTypes <- all_cellType_order
start_time = Sys.time()
allStates.mat <- data.frame()
allAbundances <- data.frame()
myD <- 1
for(myCelltype in mySelectedCellTypes){
  
  myfile <- paste('FlowSOM_allData_states_',myCelltype,'.RDa',sep ='')
  load(myfile) 
  
  #first concatenate the heatmap with all states
  heat_mat_test <- as.data.frame(heat_mat_test1)
  heat_mat_test$CellType <- myCelltype
  heat_mat_test$ID <- paste('s',1:nrow(heat_mat_test),sep='')
  heat_mat_test$ID2 <- paste(myCelltype,'_s',1:nrow(heat_mat_test),sep='')
  rownames(heat_mat_test) <- heat_mat_test$ID
  allStates.mat <- rbind(allStates.mat,heat_mat_test)
  
  #next concatenate the relative abundances
  myAbundance <- data_df_test %>% 
    group_by(Sample,pop_labels) %>% 
    summarise(samples = n()) %>%
    mutate(n=samples/sum(samples))
  
  pop_labels <- as.numeric(myAbundance$pop_labels)
  
  myAbundance <- as.data.frame(myAbundance)
  myAbundance$Sample <- factor(myAbundance$Sample,levels = c(groupA,groupD,groupE,groupF))
  myAbundance$pop_labels <- factor(myAbundance$pop_labels,levels = c(1:max(pop_labels)))
  myAbundance$ID <- paste('s',myAbundance$pop_labels,sep='') 
  myAbundance$ID2 <- paste(myCelltype,'_s',myAbundance$pop_labels,sep='')
  myAbundance$CellType <- myCelltype
  
  allAbundances <- rbind(allAbundances,myAbundance)
  
  #make one plot per cell type
  #make a heatmap of all states
  my_row_annot <- data.frame(CellType=heat_mat_test$CellType)  
  rownames(my_row_annot) <- rownames(heat_mat_test)
  
  myD <-myD + 1    
  
} #end of for loop to parse the cell types
end_time = Sys.time()
end_time-start_time


#visualise the abundance of cellular states per cell type 
allAbundances$ID <- factor(allAbundances$ID,levels =  paste('s',c(1:length(unique(allAbundances$ID))),sep='') )
allAbundances$CellType <- factor(allAbundances$CellType, levels=all_cellType_order)

myPalete <- c("#FEE6CE","#FDAE6B","#D94801","#FDE0EF","#9E9AC8", "#54278F","#C6DBEF","#4292C6", "#084594",'khaki3') 

o <- ggplot(allAbundances) + aes(x=Sample,y=n,fill=ID)+
  geom_bar(stat='identity',position='stack',width = 0.48,alpha=1,color='black',size=0.15)+
  #geom_hline(yintercept = 0.25,color='red',size=0.25)+
  #geom_hline(yintercept = 0.875,color='red',size=0.25)+
  facet_wrap(~CellType,scales='free_y',nrow=4)+
  #geom_text(aes(y = n, label = ID), position = position_stack(vjust = 0.5),color = "black", size=2)+
  theme_clean()+
  ylab('Relative abundance')+
  xlab('Distinct cellular states per cell type')+
  theme(axis.text.y = element_text( size = 8 ),
        axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 0.5, size = 8),
        axis.title.x = element_text( size = 8,face = 'bold' ),
        axis.title.y = element_text( size = 8 ),
        strip.text = element_text(size = 8,face='bold',lineheight=1),
        legend.position = "bottom",aspect.ratio = 0.8)+
  scale_fill_manual(values=myPalete)+
  guides(fill = guide_legend(override.aes = list(size=1.5),nrow=2,title=""))

myfile <- paste('plots/Figure_4unmodified.pdf',sep ='')
pdf(myfile)
print(o)
dev.off()

```

### Self consistency test to assess the performance of LDA in the discovery cohort

```{r performance in the discovery cohort, echo=TRUE, cache=TRUE,eval=FALSE}

mySelectedCellTypes <- c("CD4_Memory" ,"CD4_Naive" ,
                         "CD8_Memory","CD8_Naive",
                         "T_DN","T_other",
                         "B_Memory" ,"B_Naive" ,'B_other',
                         "NK_CD16_pos", "NK_CD16_neg", "NK_other",
                         "Monocytes_classical", "Monocytes_nonClassical", "DC","DC_plasmocytoid","Myelocytes_other")

########################################################################################################
########################################################################################################
#using standard performance metrics
performanceLDA <- data.frame()
for(myCelltype in mySelectedCellTypes){
  
  myfile <- paste('FlowSOM_allData_states_',myCelltype,'.RDa',sep ='')
  load(myfile) 
  
  IDX <- sample(nrow(data_df_train))
  Ncell <- floor(0.5*nrow(data_df_train))
  selectTrainIDX <- IDX[1:Ncell]
  selectTestIDX <- IDX[(Ncell+1):length(IDX)]
  trainingSet <- data_df_train[selectTrainIDX,]
  testingSet <- data_df_train[selectTestIDX,]
  #check the dimensions of the data
  dim(trainingSet)
  dim(testingSet)
  nrow(testingSet)+nrow(trainingSet)==nrow(data_df_train)
  #prepare the data
  testingSetLabels <- testingSet$pop_labels
  testingSetValues <- as.matrix(testingSet[,1:(ncol(testingSet)-3)])
  trainingSetLabels <- trainingSet$pop_labels
  trainingSetValues <- as.matrix(trainingSet[,1:(ncol(trainingSet)-3)])
  #APPROACH #4: LDA 
  tmp <- data.frame(trainingSetValues)
  tmp$CellType <- trainingSetLabels
  modelLDA <- lda(CellType~.,data=tmp)
  predictedLabel <- predict(modelLDA, newdata = data.frame(testingSetValues))
  #save the max posterior probability
  prob_matrix<-matrix(0L, nrow = nrow(testingSetValues), ncol =1)
  prob_matrix <- as.matrix(apply(predictedLabel$posterior,1,max))
  
  predictionsLDA <- data.frame(realLabel=testingSetLabels,
                               predictedLabel=predictedLabel$class,
                               Posterior=prob_matrix[,1])
  dt <- data.frame()
  for(idx in unique(testingSetLabels)){
    
    currentType <- idx
    classifiedAsCurrent <- predictionsLDA[predictionsLDA$predictedLabel==currentType,]
    tp_row <- length(which(classifiedAsCurrent$realLabel==currentType))
    classifiedAsNonCurrent <- predictionsLDA[predictionsLDA$predictedLabel!=currentType,]
    tn_row <- length(which(classifiedAsNonCurrent$realLabel!=currentType))
    fp_row <- length(which(classifiedAsNonCurrent$realLabel==currentType))
    fn_row <- length(which(classifiedAsCurrent$realLabel!=currentType))
    
    a <- data.frame(Sen=tp_row/(tp_row+fn_row),
                    Spe=tn_row/(tn_row+fp_row),
                    #PPV=tp_row/(tp_row+fp_row),
                    F1=2*tp_row/(2*tp_row+fp_row+fn_row),
                    Acc=(tp_row+tn_row)/(tp_row+tn_row+fp_row+fn_row),
                    State=idx)
    dt <- rbind(dt,a)
  }
  dt$CellType <- myCelltype
  performanceLDA <- rbind(performanceLDA,dt)
  
}

all_cellType_order <- c("CD4_Memory" ,"CD4_Naive" ,
                        "CD8_Memory","CD8_Naive",
                        "T_DN","T_other",
                        "B_Memory" ,"B_Naive" ,'B_other',
                        "NK_CD16_pos", "NK_CD16_neg", "NK_other",
                        "Monocytes_classical", "Monocytes_nonClassical", "DC","DC_plasmocytoid","Myelocytes_other")

color_qual_flow2 <- c('dodgerblue','coral1',"#FFD258","#3F1B03", "#894F36",'#9a6324',"#F4AD31",'khaki3',
                      "#F4800C",'darkorange3' , 'brown2',"#CC242A",'firebrick4',
                      "#EF8ECC",'#e6beff','#fabebe','darkorchid4','lightpink3','mediumpurple2',
                      "#1C750C","#557F7A","#4DB23B","#066970",'#008080',"#b0c6a2",
                      '#808000',"#CBD49C",'olivedrab2','mediumaquamarine','paleturquoise1','steelblue3',
                      "#6471E2",'skyblue','#4363d8','#46f0f0','#000075','slategray3',
                      'mediumblue','cornflowerblue','#ffe119','#ffd8b1',"#FEEDC3",
                      'bisque3','#808080','black','gray88','gray','burlywood4')

tmp <- melt(performanceLDA,id.vars = c("State","CellType" ))
tmp$CellType <- factor(tmp$CellType,levels = all_cellType_order)
colnames(tmp)[3] <- 'Metric'
tmp$Metric <- factor(tmp$Metric,levels = c('Acc','Sen','Spe','F1'))

tmp$State <- factor(tmp$State,levels=1:10)

myPalete <- c("#FEE6CE","#FDAE6B","#D94801","#FDE0EF","#9E9AC8", "#54278F","#C6DBEF","#4292C6", "#084594",'khaki3') 

performance_sanityTest <- ggplot(tmp, aes(x = Metric, y = value, color=State,fill = Metric)) +
  geom_jitter(height=0,width=0.2, size=1,alpha=0.8)+
  geom_boxplot(width=0.15,
               alpha=0.3,
               size=0.25,
               outlier.colour = "black",
               outlier.shape = NA,
               outlier.fill = "red",
               outlier.size = 0.05,
               notch = FALSE,color='black')+
  facet_wrap(~CellType,scales='free_y',nrow=4)+
  ggtitle('LDA self-consistency test - Discovery cohort')+
  theme_clean() +
  ylim(0.5,1)+
  scale_fill_manual(values = color_qual_flow2)+
  scale_color_manual(values = myPalete)+
  theme(axis.text.y = element_text( size = 12 ),
        axis.text.x = element_text(angle = 90, vjust = 0.05, hjust = 0.95, size = 12),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        strip.text = element_text(size = 12,face='bold',lineheight=1),
        legend.position = "none",aspect.ratio = 0.8)

myfile <- paste('plots/Supp_Figure_9A.pdf',sep ='')
pdf(myfile, width = 8, height = 10)
print(performance_sanityTest)
dev.off()

########################################################################################################
#using correlation as a metric of agreement of the predicted states

colors_all <- c('slategray1','slategray3',
                'steelblue1','steelblue3',
                'darkslategrey','slateblue4',
                'firebrick1','firebrick4','indianred1',
                'orange','orange3','peachpuff',
                'palegreen3','palegreen1','lightgoldenrod3','lightgoldenrod1','greenyellow')

summaryCellsStates <- data.frame()
corrAgree <- data.frame()
corrDisagree <- data.frame()

for(myCelltype in mySelectedCellTypes){
  
  myfile <- paste('/Home/siv32/dkl081/Desktop/RAMMS/paper_figure_codes/state_discovery_plots/data_files/FlowSOM_allData_states_',myCelltype,'.RDa',sep ='')
  load(myfile) 
  
  stateNumberTest <- length(unique(data_df_test$pop_labels))
  stateNumberTrain <- length(unique(data_df_train$pop_labels))
  
  myMatrix <- matrix(-1,nrow=stateNumberTest,ncol=stateNumberTrain)
  numCells <- matrix(-1,nrow=stateNumberTrain,ncol=1)
  for(i in 1:stateNumberTest){
    
    tmp_testData <- data_df_test[data_df_test$pop_labels==i,1:28]
    mean_testData <- colMeans(tmp_testData)
    
    for(j in 1:stateNumberTrain){
      
      tmp_trainData <- data_df_train[data_df_train$pop_labels==j,1:28]
      mean_trainData <- colMeans(tmp_trainData)
      myMatrix[i,j] <- cor(mean_testData,mean_trainData)
      numCells[j,1] <- nrow(tmp_trainData)
      
      #skip the cases where we have few cells
      if(nrow(tmp_trainData)>=100){
        
        if(i==j){
          a1 <- data.frame(State=j,
                           CellType=myCelltype,
                           Corr=cor(mean_testData,mean_trainData))
          corrAgree <- rbind(corrAgree,a1)
        }else{
          a2 <- data.frame(State_i=i,
                           State_j=j,
                           CellType=myCelltype,
                           Corr=cor(mean_testData,mean_trainData))
          corrDisagree <- rbind(corrDisagree,a2)
        }
      }
    }
  }
  dt <- data.frame(State=1:stateNumberTrain,
                   numCells,
                   CellType=myCelltype)
  
  summaryCellsStates <- rbind(summaryCellsStates,dt)
}

corrAgree$Status <- 'Predicted'
corrDisagree$Status <- 'nonPredicted'
colnames(corrDisagree)[2] <- 'State'

corrDisagree$State <- 0

tmp <- rbind(corrAgree[,1:4],corrDisagree[,2:5])
tmp$CellType <- factor(tmp$CellType,levels = all_cellType_order)
tmp$Status <- factor(tmp$Status,levels = c('Predicted','nonPredicted'))
tmp$State <- factor(tmp$State,levels=0:10)

myPalete <- c('gray',"#FEE6CE","#FDAE6B","#D94801","#FDE0EF","#9E9AC8", "#54278F","#C6DBEF","#4292C6", "#084594",'khaki3') 

corr_sanityTest <- ggplot(tmp, aes(x = Status, y = Corr, color=State,fill = CellType)) +
  geom_jitter(height=0,width=0.2, size=2,alpha=0.8)+
  geom_boxplot(width=0.1,
               alpha=0.3,
               size=0.25,
               outlier.colour = "black",
               outlier.shape = NA,
               outlier.fill = "red",
               outlier.size = 0.05,
               notch = FALSE,color='black')+
  facet_wrap(~CellType,scales='free_y',nrow=4)+
  theme_clean() +
  ylab('Pearsons correlation ')+
  #ylim(0.5,1)+
  scale_color_manual(values = myPalete)+
  scale_fill_manual(values = colors_all)+
  theme(axis.text.y = element_text( size = 12 ),
        axis.text.x = element_text(angle = 90, vjust = 0.05, hjust = 0.95, size = 12),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        strip.text = element_text(size = 12,face='bold',lineheight=1),
        legend.position = "none",aspect.ratio = 0.8)


myfile <- paste('plots/Supp_Figure_9B.pdf',sep ='')
pdf(myfile, width = 8, height = 10)
print(corr_sanityTest)
dev.off()

```

### Visualise the cellular states in the discovery cohort using functional graphs and generate heatmaps showing the average expression profiles of all cellular states per compartment

```{r make functional graphs and heatmaps, echo=TRUE, cache=TRUE,eval=FALSE}

combinedScaledHeatmaps <- data.frame()

#load an in-house function
source('computeCellTypeDistance.R')

###########################################################
# T cells
###########################################################
a <- which(allStates.mat$CellType=='CD4_Memory'| 
               allStates.mat$CellType=='CD4_Naive'|
               allStates.mat$CellType=='CD8_Memory'| 
               allStates.mat$CellType=='CD8_Naive'| 
               allStates.mat$CellType=='T_DN'| 
               allStates.mat$CellType=='T_other')
tmp <- allStates.mat[a,]

my_colour <- list(
  CellType=c(CD4_Memory=colors_all[1],
             CD4_Naive=colors_all[2],
             CD8_Memory=colors_all[3],
             CD8_Naive=colors_all[4],
             T_DN=colors_all[5],
             T_other=colors_all[6])
)

myfile <- paste('plots/Supp_Figure_10.pdf',sep ='')

rownames(tmp) <- tmp$ID2
tmp2 <- tmp[,1:28]
tmp2[tmp2 > quantile(as.matrix(tmp[,1:28]),0.99)] <- quantile(as.matrix(tmp[,1:28]),0.99)

breaksList <- seq(0, 1.8, by = 0.2)
pheatmap::pheatmap(tmp2[,1:28],
                   cluster_cols = F,
                   cluster_rows = F,
                   filename =myfile,
                   display_numbers=F,
                   fontsize_number = 6,
                   cutree_rows = 14,
                   fontsize_row = 10,
                   width=9,
                   height=5.5,
                   breaks = breaksList,
                   color = brewer.pal(9,'Greys'),
                   annotation_row = my_row_annot,
                   annotation_colors = my_colour,
                   gaps_row = c(8,17,27,35,38) )

combinedScaledHeatmaps <- rbind(combinedScaledHeatmaps,tmp2)
combinedScaledHeatmaps$CellType <- tmp$CellType

a <- which(allAbundances$CellType=='CD4_Memory'| 
           allAbundances$CellType=='CD4_Naive'| 
           allAbundances$CellType=='CD8_Memory'| 
           allAbundances$CellType=='CD8_Naive'| 
           allAbundances$CellType=='T_DN'| 
           allAbundances$CellType=='T_other')
tmp_abundance <- allAbundances[a,]
sum_abundance <- tmp_abundance %>% 
  group_by(ID2) %>% 
  summarise(samples = sum(samples) )

filtering_T <- 0.9
#dd_Train <- computeCellTypeDistance(tmp[,1:23],colnames(tmp[1:23]) ,filtering_T)
dd_Train <- matrix(-1,nrow = nrow(tmp),ncol = nrow(tmp))
for(idx1 in 1:nrow(tmp)){
  a <- tmp[idx1,1:28]
  for(idx2 in 1:nrow(tmp)){
    b <- tmp[idx2,1:28]
    dd_Train[idx1,idx2] <- cor(unlist(a),unlist(b))
  }
}

dd_Train[dd_Train<filtering_T] <- 0

G.training <- igraph::graph_from_adjacency_matrix(dd_Train, weighted=TRUE,mode = "undirected")

message("Running ForceAtlas2...")
flush.console()

G.training <- complete_forceatlas2(G.training, 
                                   first.iter = 500000, 
                                   overlap.method = 'repel', 
                                   overlap.iter=10000,
                                   ew.influence=10000,
                                   kgrav=18)
#obtain the vertex attributes
G.training.attr <- vertex_attr(G.training)
#the coordinates of the landmarks can be used for visualisation
G.training.attr <- data.frame(x=G.training.attr$x,
                              y=G.training.attr$y,
                              Name=tmp$ID2,
                              CellType=tmp$CellType)

sum_abundance$Flag <- ''
sum_abundance[sum_abundance$ID2=='CD4_Naive_s1'|
              sum_abundance$ID2=='CD4_Memory_s3'|
              sum_abundance$ID2=='CD8_Memory_s1'|
              sum_abundance$ID2=='CD8_Memory_s8'|
              sum_abundance$ID2=='CD8_Naive_s7'|
              sum_abundance$ID2=='CD8_Naive_s8'|  
              sum_abundance$ID2=='T_other_s3'  
                ,'Flag'] <- '<100 cells'

dr <- match(tmp$ID2,sum_abundance$ID2)
V(G.training)$Name <- factor(tmp$ID,levels = paste('s',1:length(unique(tmp$ID)),sep=''))
V(G.training)$CellType <-  factor(tmp$CellType,levels = c('CD4_Memory','CD4_Naive',
                                                          'CD8_Memory','CD8_Naive','T_DN','T_other'))
V(G.training)$Cells <- sum_abundance$Flag[dr]

o <- ggraph(G.training) +
  geom_edge_link(alpha=.4,edge_colour='black')+
  geom_node_point(aes(fill=CellType),shape=21,size=15)+
  geom_node_text(aes(label=Name),fontface='bold',size=8)+
  geom_node_text(aes(filter=Cells=='<100 cells',label='<100'),size=8,lineheight=8,colour='red4')+
  scale_fill_manual(values=colors_all[1:6])+
  scale_color_brewer(palette='Set3')+
  theme_graph(base_family = 'Helvetica') + 
  theme(legend.position="bottom",
        legend.text = element_text(size=14))+
  guides(fill = guide_legend(override.aes = list(size=14),nrow=2,title=""))

myfile <- paste('plots/Supp_Figure_14a.pdf',sep ='')
pdf(myfile,width = 20,height = 7.5)
print(o)
dev.off()

  
###########################################################
# B cells
###########################################################
a <- which(allStates.mat$CellType=='B_Memory'| 
             allStates.mat$CellType=='B_Naive' | 
             allStates.mat$CellType=='B_other')
tmp <- allStates.mat[a,]

my_colour <- list(
  CellType=c(B_Memory=colors_all[7],
             B_Naive=colors_all[8],
             B_other=colors_all[9]))

myfile <- paste('plots/Supp_Figure_11.pdf',sep ='')

rownames(tmp) <- tmp$ID2
tmp2 <- tmp[,1:28]
tmp2[tmp2 > quantile(as.matrix(tmp[,1:28]),0.99)] <- quantile(as.matrix(tmp[,1:28]),0.99)

breaksList <- seq(0, 1.8, by = 0.2)

pheatmap::pheatmap(tmp2[,1:28],
                   cluster_cols = F,
                   cluster_rows = F,
                   filename =myfile,
                   display_numbers=F,
                   fontsize_number = 6,
                   cutree_rows = 14,
                   fontsize_row = 10,
                   width=9,
                   height=5.5,
                   breaks = breaksList,
                   color = brewer.pal(9,'Greys'),
                   annotation_row = my_row_annot,
                   annotation_colors = my_colour,
                   gaps_row = c(8,13) )

tmp2$CellType <- tmp$CellType
combinedScaledHeatmaps <- rbind(combinedScaledHeatmaps,tmp2)

a <- which(allAbundances$CellType=='B_Memory'| 
             allAbundances$CellType=='B_Naive'| 
             allAbundances$CellType=='B_other' )
tmp_abundance <- allAbundances[a,]
sum_abundance <- tmp_abundance %>% 
  group_by(ID2) %>% 
  summarise(samples = sum(samples) )

filtering_T <- 0.9
#dd_Train <- computeCellTypeDistance(tmp[,1:23],colnames(tmp[1:23]) ,filtering_T)
dd_Train <- matrix(-1,nrow = nrow(tmp),ncol = nrow(tmp))
for(idx1 in 1:nrow(tmp)){
  a <- tmp[idx1,1:28]
  for(idx2 in 1:nrow(tmp)){
    b <- tmp[idx2,1:28]
    dd_Train[idx1,idx2] <- cor(unlist(a),unlist(b))
  }
}

dd_Train[dd_Train<filtering_T] <- 0

G.training <- igraph::graph_from_adjacency_matrix(dd_Train, weighted=TRUE,mode = "undirected")

message("Running ForceAtlas2...")
flush.console()
G.training <- complete_forceatlas2(G.training, 
                                   first.iter = 500000, 
                                   overlap.method = 'repel', 
                                   overlap.iter=10000,
                                   ew.influence=10000,
                                   kgrav=18)
#obtain the vertex attributes
G.training.attr <- vertex_attr(G.training)
#the coordinates of the landmarks can be used for visualisation
G.training.attr <- data.frame(x=G.training.attr$x,
                              y=G.training.attr$y,
                              Name=tmp$ID2,
                              CellType=tmp$CellType)

sum_abundance$Flag <- ''
sum_abundance[sum_abundance$ID2=='B_Memory_s3' |
                sum_abundance$ID2=='B_Naive_s3' |
                sum_abundance$ID2=='B_Naive_s5' |
                sum_abundance$ID2=='B_other_s1' |  
                sum_abundance$ID2=='B_other_s3' |
                sum_abundance$ID2=='B_other_s4' 
              ,'Flag'] <- '<100 cells'

dr <- match(tmp$ID2,sum_abundance$ID2)
V(G.training)$Name <- factor(tmp$ID,levels = paste('s',1:length(unique(tmp$ID)),sep=''))
V(G.training)$CellType <-  factor(tmp$CellType,levels = c('B_Memory','B_Naive','B_other'))
V(G.training)$Cells <- sum_abundance$Flag[dr]

o <- ggraph(G.training) +
  geom_edge_link(alpha=.4,edge_colour='black')+
  geom_node_point(aes(fill=CellType),shape=21,size=15)+
  geom_node_text(aes(label=Name),fontface='bold',size=8)+
  geom_node_text(aes(filter=Cells=='<100 cells',label='<100'),size=8,lineheight=8,colour='red4')+
  scale_fill_manual(values=colors_all[7:9])+
  scale_color_brewer(palette='Set3')+
  theme_graph(base_family = 'Helvetica') + 
  theme(legend.position="bottom",
        legend.text = element_text(size=14))+
  guides(fill = guide_legend(override.aes = list(size=14),nrow=2,title=""))

myfile <- paste('plots/Supp_Figure_14b.pdf',sep ='')
pdf(myfile,width = 20,height = 7.5)
print(o)
dev.off()      

###########################################################
# NK cells
###########################################################   
a <- which(allStates.mat$CellType=='NK_CD16_pos'| 
             allStates.mat$CellType=='NK_CD16_neg' | 
             allStates.mat$CellType=='NK_other')
tmp <- allStates.mat[a,]

my_colour <- list(
  CellType=c(NK_CD16_pos=colors_all[10],
             NK_CD16_neg=colors_all[11],
             NK_other=colors_all[12]))

myfile <- paste('plots/Supp_Figure_12.pdf',sep ='')

rownames(tmp) <- tmp$ID2
tmp2 <- tmp[,1:28]
tmp2[tmp2 > quantile(as.matrix(tmp[,1:28]),0.99)] <- quantile(as.matrix(tmp[,1:28]),0.99)

breaksList <- seq(0, 1.8, by = 0.2)

pheatmap::pheatmap(tmp2[,1:28],
                   cluster_cols = F,
                   cluster_rows = F,
                   filename =myfile,
                   display_numbers=F,
                   fontsize_number = 6,
                   cutree_rows = 14,
                   fontsize_row = 10,
                   width=9,
                   height=5.5,
                   breaks = breaksList,
                   color = brewer.pal(9,'Greys'),
                   annotation_row = my_row_annot,
                   annotation_colors = my_colour,
                   gaps_row = c(5,13) )

tmp2$CellType <- tmp$CellType
combinedScaledHeatmaps <- rbind(combinedScaledHeatmaps,tmp2)

a <- which(allAbundances$CellType=='NK_CD16_pos'| 
             allAbundances$CellType=='NK_CD16_neg'| 
             allAbundances$CellType=='NK_other' )
tmp_abundance <- allAbundances[a,]
sum_abundance <- tmp_abundance %>% 
  group_by(ID2) %>% 
  summarise(samples = sum(samples) )

filtering_T <- 0.9
#dd_Train <- computeCellTypeDistance(tmp[,1:23],colnames(tmp[1:23]) ,filtering_T)
dd_Train <- matrix(-1,nrow = nrow(tmp),ncol = nrow(tmp))
for(idx1 in 1:nrow(tmp)){
  a <- tmp[idx1,1:28]
  for(idx2 in 1:nrow(tmp)){
    b <- tmp[idx2,1:28]
    dd_Train[idx1,idx2] <- cor(unlist(a),unlist(b))
  }
}

dd_Train[dd_Train<filtering_T] <- 0

G.training <- igraph::graph_from_adjacency_matrix(dd_Train, weighted=TRUE,mode = "undirected")


message("Running ForceAtlas2...")
flush.console()
G.training <- complete_forceatlas2(G.training, 
                                   first.iter = 500000, 
                                   overlap.method = 'repel', 
                                   overlap.iter=10000,
                                   ew.influence=10000,
                                   kgrav=18)
#obtain the vertex attributes
G.training.attr <- vertex_attr(G.training)
#the coordinates of the landmarks can be used for visualisation
G.training.attr <- data.frame(x=G.training.attr$x,
                              y=G.training.attr$y,
                              Name=tmp$ID2,
                              CellType=tmp$CellType)

sum_abundance$Flag <- ''
sum_abundance[sum_abundance$ID2=='NK_CD16_neg_s3' |
                sum_abundance$ID2=='NK_CD16_neg_s8' |
                sum_abundance$ID2=='NK_other_s4' 
              ,'Flag'] <- '<100 cells'


dr <- match(tmp$ID2,sum_abundance$ID2)
V(G.training)$Name <- factor(tmp$ID,levels = paste('s',1:length(unique(tmp$ID)),sep=''))
V(G.training)$CellType <- factor(tmp$CellType,levels = c('NK_CD16_pos','NK_CD16_neg','NK_other'))
V(G.training)$Cells <- sum_abundance$Flag[dr]

o <- ggraph(G.training) +
  geom_edge_link(alpha=.4,edge_colour='black')+
  geom_node_point(aes(fill=CellType),shape=21,size=15)+
  geom_node_text(aes(label=Name),fontface='bold',size=8)+
  geom_node_text(aes(filter=Cells=='<100 cells',label='<100'),size=8,lineheight=8,colour='red4')+
  scale_fill_manual(values=colors_all[10:12])+
  scale_color_brewer(palette='Set3')+
  theme_graph(base_family = 'Helvetica') + 
  theme(legend.position="bottom",
        legend.text = element_text(size=14))+
  guides(fill = guide_legend(override.aes = list(size=14),nrow=2,title=""))

myfile <- paste('plots/Supp_Figure_14c.pdf',sep ='')
pdf(myfile,width = 20,height = 7.5)
print(o)
dev.off()        

###########################################################
# Myelocytes cells
###########################################################    

a <- which(allStates.mat$CellType=='Monocytes_classical'| 
             allStates.mat$CellType=='Monocytes_nonClassical' | 
             allStates.mat$CellType=='DC'| 
             allStates.mat$CellType=='DC_plasmocytoid'| 
             allStates.mat$CellType=='Myelocytes_other')
tmp <- allStates.mat[a,]

my_colour <- list(
  CellType=c(Monocytes_classical=colors_all[13],
             Monocytes_nonClassical=colors_all[14],
             DC=colors_all[15],
             DC_plasmocytoid=colors_all[16],
             Myelocytes_other=colors_all[17])
)

myfile <- paste('plots/Supp_Figure_13.pdf',sep ='')

rownames(tmp) <- tmp$ID2
tmp2 <- tmp[,1:28]
tmp2[tmp2 > quantile(as.matrix(tmp[,1:28]),0.966)] <- quantile(as.matrix(tmp[,1:28]),0.966)

breaksList <- seq(0, 1.8, by = 0.2)

pheatmap::pheatmap(tmp2[,1:28],
                   cluster_cols = F,
                   cluster_rows = F,
                   filename =myfile,
                   display_numbers=F,
                   fontsize_number = 6,
                   cutree_rows = 14,
                   fontsize_row = 10,
                   width=9,
                   height=5.5,
                   breaks = breaksList,
                   color = brewer.pal(9,'Greys'),
                   annotation_row = my_row_annot,
                   annotation_colors = my_colour,
                   gaps_row = c(7,15,19,24) )

tmp2$CellType <- tmp$CellType
combinedScaledHeatmaps <- rbind(combinedScaledHeatmaps,tmp2)

a <- which(allAbundances$CellType=='Monocytes_classical'| 
             allAbundances$CellType=='Monocytes_nonClassical'| 
             allAbundances$CellType=='DC'| 
             allAbundances$CellType=='DC_plasmocytoid'| 
             allAbundances$CellType=='Myelocytes_other')
tmp_abundance <- allAbundances[a,]
sum_abundance <- tmp_abundance %>% 
  group_by(ID2) %>% 
  summarise(samples = sum(samples) )

filtering_T <- 0.9
#dd_Train <- computeCellTypeDistance(tmp[,1:23],colnames(tmp[1:23]) ,filtering_T)
dd_Train <- matrix(-1,nrow = nrow(tmp),ncol = nrow(tmp))
for(idx1 in 1:nrow(tmp)){
  a <- tmp[idx1,1:28]
  for(idx2 in 1:nrow(tmp)){
    b <- tmp[idx2,1:28]
    dd_Train[idx1,idx2] <- cor(unlist(a),unlist(b))
  }
}

dd_Train[dd_Train<filtering_T] <- 0

G.training <- igraph::graph_from_adjacency_matrix(dd_Train, weighted=TRUE,mode = "undirected")
message("Running ForceAtlas2...")
flush.console()
G.training <- complete_forceatlas2(G.training, 
                                   first.iter = 500000, 
                                   overlap.method = 'repel', 
                                   overlap.iter=10000,
                                   ew.influence=10000,
                                   kgrav=18)
#obtain the vertex attributes
G.training.attr <- vertex_attr(G.training)
#the coordinates of the landmarks can be used for visualisation
G.training.attr <- data.frame(x=G.training.attr$x,
                              y=G.training.attr$y,
                              Name=tmp$ID2,
                              CellType=tmp$CellType)

sum_abundance$Flag <- ''
sum_abundance[sum_abundance$ID2=='Monocytes_classical_s5' |
                sum_abundance$ID2=='Monocytes_nonClassical_s1' |
                sum_abundance$ID2=='Monocytes_nonClassical_s3' |
                sum_abundance$ID2=='Monocytes_nonClassical_s7' |  
                sum_abundance$ID2=='DC_s2' |
                sum_abundance$ID2=='DC_s4' |
                sum_abundance$ID2=='DC_plasmocytoid_s1' |
                sum_abundance$ID2=='Myelocytes_other_s1'|
                sum_abundance$ID2=='Myelocytes_other_s3'|
                sum_abundance$ID2=='Myelocytes_other_s4'|
                sum_abundance$ID2=='Myelocytes_other_s5'          
              ,'Flag'] <- '<100 cells'

dr <- match(tmp$ID2,sum_abundance$ID2)
V(G.training)$Name <- factor(tmp$ID,levels = paste('s',1:length(unique(tmp$ID)),sep=''))
V(G.training)$CellType <- factor(tmp$CellType,levels = c('Monocytes_classical','Monocytes_nonClassical',
                                                         'DC','DC_plasmocytoid','Myelocytes_other'))
V(G.training)$Cells <- sum_abundance$Flag[dr]

o <- ggraph(G.training) +
  geom_edge_link(alpha=.4,edge_colour='black')+
  geom_node_point(aes(fill=CellType),shape=21,size=15)+
  geom_node_text(aes(label=Name),fontface='bold',size=8)+
  geom_node_text(aes(filter=Cells=='<100 cells',label='<100'),size=8,lineheight=8,colour='red4')+
  scale_fill_manual(values=colors_all[13:17])+
  scale_color_brewer(palette='Set3')+
  theme_graph(base_family = 'Helvetica') + 
  theme(legend.position="bottom",
        legend.text = element_text(size=18))+
  guides(fill = guide_legend(override.aes = list(size=14),nrow=2,title=""))

myfile <- paste('plots/Supp_Figure_14d.pdf',sep ='')
pdf(myfile,width = 20,height = 7.5)
print(o)
dev.off()        


####################################################################################

#make a heatmap of all states
my_row_annot <- data.frame(CellType=combinedScaledHeatmaps$CellType)
rownames(my_row_annot) <- rownames(combinedScaledHeatmaps)

my_colour <- list(
  CellType=c(CD4_Memory=colors_all[1],
             CD4_Naive=colors_all[2],
             CD8_Memory=colors_all[3],
             CD8_Naive=colors_all[4],
             T_DN=colors_all[5],
             T_other=colors_all[6],
             B_Memory=colors_all[7],
             B_Naive=colors_all[8],
             B_other=colors_all[9],
             NK_CD16_pos=colors_all[10],
             NK_CD16_neg=colors_all[11],
             NK_other=colors_all[12],
             Monocytes_classical=colors_all[13],
             Monocytes_nonClassical=colors_all[14],
             DC=colors_all[15],
             DC_plasmocytoid=colors_all[16],
             Myelocytes_other=colors_all[17]
  )
)


myfile <- paste('/Home/siv32/dkl081/Desktop/RAMMS/paper_figure_codes/state_discovery_plots/cellular_states/allStates_heatmap.pdf',sep ='')

pheatmap::pheatmap(combinedScaledHeatmaps[,1:28],
                   cluster_cols = F,
                   cluster_rows = F,
                   filename =myfile,
                   display_numbers=F,
                   fontsize_number = 4,
                   cutree_rows = 14,
                   fontsize_row = 10,
                   width = 8,
                   height = 21,
                   annotation_row = my_row_annot,
                   annotation_colors = my_colour,
                   color = brewer.pal(9,'Greys'),
                   gaps_row = c(8,17,27,35,38,42,50,55,60,65,73,77,84,92,96,101) )  
```

### Compute fold changes in the abundance of cellular states within the discovery cohort.
We use BL versus all time points up to 12 months after aHSCT

```{r fold changes discovery cohort, echo=TRUE, cache=TRUE,eval=FALSE}

#compute the fold changes first for the time points of the discovery cohort
abundance_V4 <- allAbundances[allAbundances$Sample=='BL',]
abundance_V5 <- allAbundances[allAbundances$Sample=='100d',]
abundance_V6 <- allAbundances[allAbundances$Sample=='6m',]
abundance_V7 <- allAbundances[allAbundances$Sample=='12m',]
#concatenate the data
tmp <- merge(abundance_V4,abundance_V5,by='ID2')
tmp <- tmp[,c(1,5,11)]
colnames(tmp)[2] <- 'V4'
colnames(tmp)[3] <- 'V5'
tmp <- merge(tmp,abundance_V6,by='ID2')
tmp <- tmp[,c(1,2,3,7)]
colnames(tmp)[4] <- 'V6'
tmp <- merge(tmp,abundance_V7,by='ID2')
tmp <- tmp[,c(1,2,3,4,8,9,10)]
colnames(tmp)[5] <- 'V7'
#compute fold changes
foldChanges <- data.frame()
for(idx in 1:nrow(tmp)){
  
  currentState <- tmp[idx,1]
  valueV4 <- tmp[idx,2]
  valueV5 <- tmp[idx,3]
  valueV6 <- tmp[idx,4]
  valueV7 <- tmp[idx,5]
  ID <- tmp[idx,6]
  cellType <- tmp[idx,7]
  
  dt <- data.frame(Celltype=cellType,
                   ID=ID,
                   CurrentState=currentState,
                   LogV4V5=log2(valueV5/valueV4),
                   LogV4V6=log2(valueV6/valueV4),
                   LogV4V7=log2(valueV7/valueV4))
  foldChanges <- rbind(foldChanges,dt)
}

#######################################################################################3
#visualise the changes

all_cellType_order <- c("CD4_Memory" ,"CD4_Naive" ,
                        "CD8_Memory","CD8_Naive",
                        "T_DN","T_other",
                        "B_Memory" ,"B_Naive" ,'B_other',
                        "NK_CD16_pos", "NK_CD16_neg", "NK_other",
                        "Monocytes_classical", "Monocytes_nonClassical", "DC","DC_plasmocytoid","Myelocytes_other")

colors_all <- c('slategray1','slategray3',
                'steelblue1','steelblue3',
                'darkslategrey','slateblue4',
                'firebrick1','firebrick4','indianred1',
                'orange','orange3','peachpuff',
                'palegreen3','palegreen1','lightgoldenrod3','lightgoldenrod1','greenyellow')

foldChanges$Celltype <- factor(foldChanges$Celltype,levels = all_cellType_order)

o1 <- ggplot(foldChanges) + aes(x=CurrentState,y=LogV4V5,color=Celltype,fill=Celltype,label=ID)+
   geom_hline(yintercept = -1.25,color='red',size=0.25,linetype='dashed')+
    geom_hline(yintercept = 1.97,color='red',size=0.25,linetype='dashed')+
    geom_hline(yintercept = 0,color='black',size=0.25,linetype='dashed')+
    geom_point(size=3.5)+
    geom_text(aes(label=ID),position = position_identity(),vjust = -1,size=3,color='black')+
    #geom_line(group=1,color='gray44',size=0.15)+
    theme_clean()+
    xlab('')+
    ylab('Fold changes 100d / BL (log2 scale)')+
    scale_color_manual(values=colors_all)+
    scale_fill_manual(values=colors_all)+
    theme(axis.text.y = element_text( size = 10,face = 'bold' ),
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.95, size = 8,face = 'bold'),
          axis.title.x = element_text( size = 8,face = 'bold' ),
          axis.title.y = element_text( size = 8,face = 'bold' ),
          strip.text = element_text(size = 8,face='bold',lineheight=1),
          legend.position = "none",aspect.ratio = 0.25)
#ordered plot
o1b <- ggplot(foldChanges) + aes(x=reorder(CurrentState,-LogV4V5),y=LogV4V5,color=Celltype,fill=Celltype,label=ID)+
   geom_hline(yintercept = -2,color='red',size=0.25,linetype='dashed')+
    geom_hline(yintercept = 2,color='red',size=0.25,linetype='dashed')+
    geom_hline(yintercept = 0,color='black',size=0.25,linetype='dashed')+
    geom_point(size=4.5)+
    geom_text(aes(label=ID),position = position_identity(),vjust = -1,size=3,color='black')+
    #geom_line(group=1,color='gray44',size=0.15)+
    theme_clean()+
    xlab('')+
    ylim(-6,6)+
    ylab('Fold changes 100d / BL (log2 scale)')+
    scale_color_manual(values=colors_all)+
    scale_fill_manual(values=colors_all)+
    theme(axis.text.y = element_text( size = 12,face = 'bold' ),
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.95, size = 12,face = 'bold'),
          axis.title.x = element_text( size = 12,face = 'bold' ),
          axis.title.y = element_text( size = 12,face = 'bold' ),
          strip.text = element_text(size = 12,face='bold',lineheight=1),
          legend.position = "none",aspect.ratio = 0.25)

myfile <- paste('plots/Supp_Figure_15a.pdf',sep ='')
pdf(myfile,width = 20,height = 7.5)
print(o1b)
dev.off()

o2 <- ggplot(foldChanges) + aes(x=CurrentState,y=LogV4V6,color=Celltype,fill=Celltype,label=ID)+
    geom_hline(yintercept = -1.25,color='red',size=0.25,linetype='dashed')+
    geom_hline(yintercept = 1.97,color='red',size=0.25,linetype='dashed')+
    geom_hline(yintercept = 0,color='black',size=0.25,linetype='dashed')+
    geom_point(size=3.5)+
    geom_text(aes(label=ID),position = position_identity(),vjust = -1,size=3,color='black')+
    #geom_line(group=1,color='gray44',size=0.15)+
    theme_clean()+
    xlab('')+
    ylab('Fold changes 6m / BL (log2 scale)')+
    scale_color_manual(values=colors_all)+
    scale_fill_manual(values=colors_all)+
    theme(axis.text.y = element_text( size = 10,face = 'bold' ),
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.95, size = 10,face = 'bold'),
          axis.title.x = element_text( size = 10,face = 'bold' ),
          axis.title.y = element_text( size = 10,face = 'bold' ),
          strip.text = element_text(size = 10,face='bold',lineheight=1),
          legend.position = "none",aspect.ratio = 0.25)

o2b <- ggplot(foldChanges) + aes(x=reorder(CurrentState,-LogV4V6),y=LogV4V6,color=Celltype,fill=Celltype,label=ID)+
    geom_hline(yintercept = -2,color='red',size=0.25,linetype='dashed')+
    geom_hline(yintercept = 2,color='red',size=0.25,linetype='dashed')+
    geom_hline(yintercept = 0,color='black',size=0.25,linetype='dashed')+
    geom_point(size=4.5)+
    geom_text(aes(label=ID),position = position_identity(),vjust = -1,size=3,color='black')+
    #geom_line(group=1,color='gray44',size=0.15)+
    theme_clean()+
    xlab('')+
    ylim(-6,6)+
    ylab('Fold changes 6m / BL (log2 scale)')+
    scale_color_manual(values=colors_all)+
    scale_fill_manual(values=colors_all)+
    theme(axis.text.y = element_text( size = 12,face = 'bold' ),
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.95, size = 12,face = 'bold'),
          axis.title.x = element_text( size = 12,face = 'bold' ),
          axis.title.y = element_text( size = 12,face = 'bold' ),
          strip.text = element_text(size = 12,face='bold',lineheight=1),
          legend.position = "none",aspect.ratio = 0.25)

myfile <- paste('plots/Supp_Figure_15b.pdf',sep ='')
pdf(myfile,width = 20,height = 7.5)
print(o2b)
dev.off()

o3 <- ggplot(foldChanges) + aes(x=CurrentState,y=LogV4V7,color=Celltype,fill=Celltype,label=ID)+
    geom_hline(yintercept = -1.25,color='red',size=0.25,linetype='dashed')+
    geom_hline(yintercept = 1.97,color='red',size=0.25,linetype='dashed')+
    geom_hline(yintercept = 0,color='black',size=0.25,linetype='dashed')+
    geom_point(size=3.5)+
    geom_text(aes(label=ID),position = position_identity(),vjust = -1,size=3,color='black')+
    #geom_line(group=1,color='gray44',size=0.15)+
    theme_clean()+
    xlab('')+
    ylab('Fold changes 12m / BL (log2 scale)')+
    scale_color_manual(values=colors_all)+
    scale_fill_manual(values=colors_all)+
    theme(axis.text.y = element_text( size = 10,face = 'bold' ),
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.95, size = 10,face = 'bold'),
          axis.title.x = element_text( size = 10,face = 'bold' ),
          axis.title.y = element_text( size = 10,face = 'bold' ),
          strip.text = element_text(size = 8,face='bold',lineheight=1),
          legend.position = "none",aspect.ratio = 0.25)

o3b <- ggplot(foldChanges) + aes(x=reorder(CurrentState,-LogV4V7),y=LogV4V7,color=Celltype,fill=Celltype,label=ID)+
    geom_hline(yintercept = -1.25,color='red',size=0.25,linetype='dashed')+
    geom_hline(yintercept = 1.97,color='red',size=0.25,linetype='dashed')+
    geom_hline(yintercept = 0,color='black',size=0.25,linetype='dashed')+
    geom_point(size=4.5)+
    geom_text(aes(label=ID),position = position_identity(),vjust = -1,size=3,color='black')+
    #geom_line(group=1,color='gray44',size=0.15)+
    theme_clean()+
    xlab('')+
    ylim(-6,6)+
    ylab('Fold changes 12m / BL (log2 scale)')+
    scale_color_manual(values=colors_all)+
    scale_fill_manual(values=colors_all)+
    theme(axis.text.y = element_text( size = 12,face = 'bold' ),
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.95, size = 12,face = 'bold'),
          axis.title.x = element_text( size = 12,face = 'bold' ),
          axis.title.y = element_text( size = 12,face = 'bold' ),
          strip.text = element_text(size = 12,face='bold',lineheight=1),
          legend.position = "none",aspect.ratio = 0.25)

myfile <- paste('plots/Supp_Figure_15c.pdf',sep ='')
pdf(myfile,width = 20,height = 7.5)
print(o3b)
dev.off()  

#estimate the 10th and 90th quantiles over all fold changes from the comparisons --> this gives us the cutoffs for filtering
dt <- c(foldChanges$LogV4V5,foldChanges$LogV4V6,foldChanges$LogV4V7)
quantile(dt,probs = seq(0,1,0.05)) 

#visualise all fold changes
colnames(foldChanges)
colnames(foldChanges)[4] <- '100d/BL'
colnames(foldChanges)[5] <- '6m/BL'
colnames(foldChanges)[6] <- '12m/BL'

plotData <- data.frame(cellType=foldChanges$Celltype,
                       currentState=foldChanges$CurrentState,
                       BL=0,
                       `100d`= foldChanges$`100d/BL`,
                       `6m`= foldChanges$`6m/BL`,
                       `12d`= foldChanges$`12m/BL`)

#make summary plots to see things 
myD <- 1
for(myCellType in mySelectedCellTypes){
  
  tmp <- melt(plotData,id.vars = c('cellType','currentState'))
  tmp <- tmp[tmp$cellType==myCellType,]
  o1 <- ggplot(tmp) + aes(x=variable,y=value,fill=cellType,color=cellType)+
    geom_hline(yintercept = -2,color='red',size=0.15,linetype='dashed')+
    geom_hline(yintercept = 2,color='red',size=0.15,linetype='dashed')+
    geom_point(size=1.5)+
    geom_line(group=1,color='black',size=0.1)+
    theme_bw()+
    xlab('')+
    ylab('Relative changes from BL (log2 scale)')+
    facet_wrap(~currentState,scales = 'fixed')+
    scale_color_manual(values=colors_all[myD])+
    theme(axis.text.y = element_text( size = 8 ),
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.95, size = 8),
          axis.title.x = element_text( size = 8,face = 'bold' ),
          axis.title.y = element_text( size = 8 ),
          strip.text = element_text(size = 8,face='bold',lineheight=1),
          legend.position = "none",aspect.ratio = 0.5)
  
  myD <- myD + 1
  
}

#threshold for the lower 
c1 <- quantile(dt,probs = seq(0,1,0.05))[3]
#threshold for the higher 
c2 <- quantile(dt,probs = seq(0,1,0.05))[19]

#simple filtering 
allFlags <- data.frame()
for(idx in 1:nrow(foldChanges)){
  flag <- 0
  tmp <- foldChanges[idx,]
  if( (tmp$`100d/BL` >=c2 | tmp$`100d/BL` <=c1) &  (tmp$`6m/BL` >=c2 | tmp$`6m/BL` <=c1) |
      (tmp$`100d/BL` >=c2 | tmp$`100d/BL` <=c1) &  (tmp$`12m/BL` >=c2 | tmp$`12m/BL` <=c1) |
      (tmp$`12m/BL` >=c2 | tmp$`12m/BL` <=c1) &  (tmp$`6m/BL` >=c2 | tmp$`6m/BL` <=c1)){
    flag <- 1
  }
  dt <- data.frame(Flag=flag)
  allFlags <- rbind(allFlags,dt)
}
foldChanges$Flag <- allFlags$Flag

filterStates <- foldChanges[foldChanges$Flag==1,]
#filter based on the number of cells from before <100 are excluded --> hard coded but i had to double check it manually
fewCellsVector <- c('CD4_Naive_s1','CD4_Memory_s3','CD8_Memory_s1','CD8_Memory_s8','CD8_Naive_s7','CD8_Naive_s8','T_other_s3','B_Memory_s3','B_Naive_s3','B_Naive_s5','B_other_s1','B_other_s3','B_other_s4',
                    'NK_CD16_neg_s3' ,'NK_CD16_neg_s8','NK_other_s4' ,'Monocytes_classical_s5','Monocytes_nonClassical_s1','Monocytes_nonClassical_s3','Monocytes_nonClassical_s7',
                    'DC_s2','DC_s4','DC_plasmocytoid_s1' ,'Myelocytes_other_s1','Myelocytes_other_s3','Myelocytes_other_s4','Myelocytes_other_s5')

a <- which(filterStates$CurrentState %in% fewCellsVector)
filterStates <- filterStates[-a,]


plotData <- data.frame(cellType=filterStates$Celltype,
                       currentState=filterStates$CurrentState,
                        BL=0,
                       `t100d`= filterStates$`100d/BL`,
                       `t6m`= filterStates$`6m/BL`,
                       `t12d`= filterStates$`12m/BL`)

tmp <- melt(plotData,id.vars = c('cellType','currentState'))

tmp$cellType <- factor(tmp$cellType,levels = c('CD4_Memory','CD4_Naive',
                                               'CD8_Memory','CD8_Naive',
                                               'T_DN','T_other',
                                               'B_Memory','B_Naive'))

tmp$currentState <- factor(tmp$currentState,levels=c('CD4_Memory_s7',
                                                     'CD4_Naive_s2','CD4_Naive_s3','CD4_Naive_s4','CD4_Naive_s6','CD4_Naive_s9',
                                                     'CD8_Memory_s2','CD8_Memory_s6','CD8_Naive_s2','CD8_Naive_s3','CD8_Naive_s4',
                                                     'CD8_Naive_s6','T_DN_s3','T_other_s4','B_Memory_s6','B_Memory_s8','B_Naive_s1'))

o1 <- ggplot(tmp) + aes(x=variable,y=value,fill=cellType,color=cellType)+
    geom_hline(yintercept = -1.25,color='red',size=0.15,linetype='dashed')+
    geom_hline(yintercept = 1.97,color='red',size=0.15,linetype='dashed')+
    geom_point(size=2.4)+
  geom_text(aes(label=round(value,1)),position = position_identity(),vjust = -1,size=3,color='black')+
    geom_line(group=1,color='black',size=0.1)+
     facet_wrap(~currentState,ncol=6,scales = 'fixed')+
     theme_clean() +
    xlab('')+
  ylim(-6,+6)+
    ylab('Relative changes from BL (log2 scale)')+
    scale_fill_manual(values = colors_all[1:8])+
    scale_color_manual(values = colors_all[1:8])+
    theme(axis.text.y = element_text( size = 8 ),
          axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 0.5, size = 8),
          axis.title.x = element_blank(),
          axis.title.y = element_text( size = 8 ),
          strip.text = element_text(size = 10,face='bold',lineheight=1),
          legend.position = "none",aspect.ratio = 0.8)

  myfile <- paste('plots/Figure_5A.pdf',sep ='')
  pdf(myfile,width = 10,height = 8)
  print(o1)
  dev.off()

```

### Predict and estimate cellular state abundances in unseen data
Here we use the validation cohort that is totally unseen from the discovery cohort. We predict the cellular states found in the discovery cohort using LDA similar to what we did for the discovery cohort.

```{r screen functional states batch2 , echo=TRUE, cache=TRUE,eval=FALSE}
#colors for time points of the discovery cohort 
color_TimePoints <- c('gray48',
                      "rosybrown4"
)  

#load the cells from the validation cohort
load('allLymphocytes_annotated.RDa')
#take out the rest of cells from the discovery cohort
allCells <- allCells[allCells$Sample=='24m'| allCells$Sample=='HC', ]

all_cellType_order <- c("CD4_Memory" ,"CD4_Naive" ,
                        "CD8_Memory","CD8_Naive",
                        "T_DN","T_other",
                        "B_Memory" ,"B_Naive" ,'B_other',
                        "NK_CD16_pos", "NK_CD16_neg", "NK_other",
                        "Monocytes_classical", "Monocytes_nonClassical", "DC","DC_plasmocytoid","Myelocytes_other")
mySelectedCellTypes <- all_cellType_order
colors_all <- c('slategray1','slategray3',
                'steelblue1','steelblue3',
                'darkslategrey','slateblue4',
                'firebrick1','firebrick4','indianred1',
                'orange','orange3','peachpuff',
                'palegreen3','palegreen1','lightgoldenrod3','lightgoldenrod1','greenyellow')

#make a small change in the name of the markers otherwise the code will crush...
colnames(allCells)[23] <- 'HLADR'
timePointsOfInterest <-  c('HC','24m')

allAbundances_validation <-data.frame()
allStates.mat <- data.frame()
myFigureList <- list()

summaryCellsStates <- data.frame()
corrAgree <- data.frame()
corrDisagree <- data.frame()

myC <- 1
for(idx in 1:length(mySelectedCellTypes)){
  
  idx
  myCelltype <- mySelectedCellTypes[idx]
   myfile <- paste('FlowSOM_allData_states_',myCelltype,'.RDa',sep ='')
  load(myfile) 
  
  myMarkers <- colnames(data_df_test)[1:28]
  
  #prepare the data for modelling using the lda.model
  myCellTypeData <- allCells[allCells$CellType==myCelltype, ]
  
  A <- predict(lda.model1,myCellTypeData[,myMarkers])
  myCellTypeData$pop_labels <- A$class
  posteriors <- as.matrix(apply(A$posterior,1,max))
  plot(density(posteriors))
  #filter the data based on prob
  idx1 <- which(posteriors>0.6)
  myCellTypeData <- myCellTypeData[idx1,]
  myCellTypeData$ID  <- paste(myCelltype,'_s',myCellTypeData$pop_labels,sep='') 
  
  #compute the correlation with the training set
  #######################################################################################
  stateNumberTest <- length(unique(myCellTypeData$pop_labels))
  stateNumberTrain <- length(unique(data_df_train$pop_labels))

  myMarkers <- colnames(data_df_train)[c(1:28)]

  myMatrix <- matrix(-1,nrow=stateNumberTest,ncol=stateNumberTrain)
  numCells <- matrix(-1,nrow=stateNumberTrain,ncol=1)
  for(i in 1:stateNumberTest){
  
    tmp_testData <- myCellTypeData[myCellTypeData$pop_labels==i,myMarkers]
    mean_testData <- colMeans(tmp_testData)
    for(j in 1:stateNumberTrain){
        tmp_trainData <- data_df_train[data_df_train$pop_labels==j,1:28]
        mean_trainData <- colMeans(tmp_trainData)
        myMatrix[i,j] <- cor(mean_testData,mean_trainData)
        numCells[j,1] <- nrow(tmp_trainData)
        #skip the cases where we have few cells
        if(nrow(tmp_trainData)>=100){
      
          if(i==j){
              a1 <- data.frame(State=j,
                         CellType=myCelltype,
                         Corr=cor(mean_testData,mean_trainData))
              corrAgree <- rbind(corrAgree,a1)
          }else{
              a2 <- data.frame(State_i=i,
                         State_j=j,
                         CellType=myCelltype,
                         Corr=cor(mean_testData,mean_trainData))
              corrDisagree <- rbind(corrDisagree,a2)
          }
      }
    }
  }
      dt <- data.frame(State=1:stateNumberTrain,
                 numCells,
                 CellType=myCelltype)

      summaryCellsStates <- rbind(summaryCellsStates,dt)
  
  #######################################################################################
  A <- as.numeric(A$class)
  ## generate a heatmap showing the expression for each cluster
  heat_mat_test <- matrix(NA, nrow = max(A), ncol = length(myMarkers))
  for(i in 1: max(A)) {
    a <- which(myCellTypeData$pop_labels==i)
    temp_mat <- myCellTypeData[a,myMarkers]
    heat_mat_test[i,] <- as.matrix(apply(temp_mat, 2, function (x) mean(x, na.rm = T)))
  }
  rownames(heat_mat_test) <- paste("cluster", 1:max(A), sep = "")
  colnames(heat_mat_test) <- myMarkers
  
  #first concatenate the heatmap with all states
  heat_mat_test <- as.data.frame(heat_mat_test)
  heat_mat_test$CellType <- myCelltype
  heat_mat_test$ID <- paste(myCelltype,'_s',1:nrow(heat_mat_test),sep='')
  rownames(heat_mat_test) <- heat_mat_test$ID
  allStates.mat <- rbind(allStates.mat,heat_mat_test)
  
  myAbundance <- myCellTypeData %>% 
    group_by(Sample,pop_labels) %>% 
    summarise(samples = n()) %>%
    mutate(n=samples/sum(samples))
  
  pop_labels <- as.numeric(myAbundance$pop_labels)
  
  myAbundance <- as.data.frame(myAbundance)
  myAbundance$Sample <- factor(myAbundance$Sample,levels = timePointsOfInterest)
  myAbundance$pop_labels <- factor(myAbundance$pop_labels,levels = c(1:max(pop_labels)))
  myAbundance$ID <- paste('s',myAbundance$pop_labels,sep='') 
  myAbundance$ID2 <- paste(myCelltype,'_s',myAbundance$pop_labels,sep='')
  myAbundance$CellType <- myCelltype
  allAbundances_validation <- rbind(allAbundances_validation,myAbundance)
  myC <- myC + 1
  
}

corrAgree$Status <- 'Predicted'
corrDisagree$Status <- 'nonPredicted'
colnames(corrDisagree)[2] <- 'State'

corrDisagree$State <- 0

tmp <- rbind(corrAgree[,1:4],corrDisagree[,2:5])
tmp$CellType <- factor(tmp$CellType,levels = all_cellType_order)
tmp$Status <- factor(tmp$Status,levels = c('Predicted','nonPredicted'))
tmp$State <- factor(tmp$State,levels=0:10)

myPalete <- c('gray',"#FEE6CE","#FDAE6B","#D94801","#FDE0EF","#9E9AC8", "#54278F","#C6DBEF","#4292C6", "#084594",'khaki3') 


corr_sanityTest <- ggplot(tmp, aes(x = Status, y = Corr, color=State,fill = CellType)) +
  geom_jitter(height=0,width=0.2, size=2,alpha=0.8)+
  geom_boxplot(width=0.1,
               alpha=0.3,
               size=0.25,
               outlier.colour = "black",
               outlier.shape = NA,
               outlier.fill = "red",
               outlier.size = 0.05,
               notch = FALSE,color='black')+
  facet_wrap(~CellType,scales='free_y',nrow=4)+
  theme_clean() +
  #ylim(0.5,1)+
  scale_color_manual(values = myPalete)+
  scale_fill_manual(values = colors_all)+
  theme(axis.text.y = element_text( size = 12 ),
        axis.text.x = element_text(angle = 90, vjust = 0.05, hjust = 0.95, size = 12),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        strip.text = element_text(size = 12,face='bold',lineheight=1),
        legend.position = "none",aspect.ratio = 0.8)

myfile <- paste('plots/Supp_Figure_16.pdf',sep ='')
pdf(myfile, width = 8, height = 10)
print(corr_sanityTest)
dev.off()

tmp <- allAbundances[allAbundances$Sample=='BL',]

plotData <- rbind(allAbundances_validation,tmp)
plotData$CellType <- factor(plotData$CellType,levels=all_cellType_order)
plotData$Sample <- factor(plotData$Sample,levels = c('BL','HC','24m'))

myPalete <- c("#FEE6CE","#FDAE6B","#D94801","#FDE0EF","#9E9AC8", "#54278F","#C6DBEF","#4292C6", "#084594",'khaki3') 
#visualise the abundance of cellular states per cell type 
plotData$ID <- factor(plotData$ID,levels =  paste('s',c(1:length(unique(plotData$ID))),sep='') )
plotData$CellType <- factor(plotData$CellType, levels=all_cellType_order)

o <- ggplot(plotData) + aes(x=Sample,y=n,fill=ID)+
  geom_bar(stat='identity',position='stack',width = 0.48,alpha=1,color='black',size=0.15)+
  #geom_hline(yintercept = 0.25,color='red',size=0.25)+
  #geom_hline(yintercept = 0.875,color='red',size=0.25)+
  facet_wrap(~CellType,scales='free_y',nrow=4)+
  #geom_text(aes(y = n, label = ID), position = position_stack(vjust = 0.5),color = "black", size=2)+
  theme_clean()+
  ylab('Relative abundance')+
  xlab('Distinct cellular states per cell type')+
  theme(axis.text.y = element_text( size = 8 ),
        axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 0.5, size = 8),
        axis.title.x = element_text( size = 8,face = 'bold' ),
        axis.title.y = element_text( size = 8 ),
        strip.text = element_text(size = 8,face='bold',lineheight=1),
        legend.position = "bottom",aspect.ratio = 0.8)+
  scale_fill_manual(values=myPalete)+
  guides(fill = guide_legend(override.aes = list(size=1.5),nrow=2,title=""))

myfile <- paste('plots/Figure_5B.pdf',sep ='')
pdf(myfile)
print(o)
dev.off()

```

## Summarise the results from the validation cohort and generate some plots

```{r combine the new data and visualise, echo=TRUE, cache=TRUE,eval=FALSE}

abundance_24m <- allAbundances_validation[allAbundances_validation$Sample=='24m',]
abundance_HC <- allAbundances_validation[allAbundances_validation$Sample=='HC',]

abundance_24m_cells <- data.frame(Sample=abundance_24m$Sample,
                                  Cells=abundance_24m$samples,
                                  ID2=abundance_24m$ID2,
                                  CellType=abundance_24m$CellType)

abundance_HC_cells <- data.frame(Sample=abundance_HC$Sample,
                                  Cells=abundance_HC$samples,
                                  ID2=abundance_HC$ID2,
                                  CellType=abundance_HC$CellType)

#concatenate the number of cells
tmp <- merge(abundance_V4,abundance_24m_cells,by='ID2')
tmp <- tmp[,c(1,4,9)]
colnames(tmp)[2] <- 'BL'
colnames(tmp)[3] <- '24m'

tmp <- merge(tmp,abundance_HC_cells,by='ID2')
tmp <- tmp[,c(1,2,3,5)]
colnames(tmp)[4] <- 'HC'

numberOfCells_part2 <- tmp

#concatenate the ratios
tmp <- merge(abundance_V4,abundance_24m,by='ID2')
tmp <- tmp[,c(1,5,11)]
colnames(tmp)[2] <- 'BL'
colnames(tmp)[3] <- '24m'

tmp <- merge(tmp,abundance_HC,by='ID2')
tmp <- tmp[,c(1,2,3,7,8,9)]
colnames(tmp)[4] <- 'HC'

#compute fold changes over the ratios
foldChanges_part2 <- data.frame()
for(idx in 1:nrow(tmp)){
  
  currentState <- tmp[idx,1]
  valueV4 <- tmp[idx,2]
  value24m <- tmp[idx,3]
  valueHC <- tmp[idx,4]
  ID <- tmp[idx,5]
  cellType <- tmp[idx,6]
  
  dt <- data.frame(Celltype=cellType,
                   ID=ID,
                   CurrentState=currentState,
                   LogV4m24=log2(value24m/valueV4),
                   LogV4HC=log2(valueHC/valueV4))
  foldChanges_part2 <- rbind(foldChanges_part2,dt)
}

foldChanges_part2$Celltype <- factor(foldChanges_part2$Celltype,levels = all_cellType_order)

o1 <- ggplot(foldChanges_part2) + aes(x=CurrentState,y=LogV4m24,color=Celltype,fill=Celltype,label=ID)+
   geom_hline(yintercept = -1.25,color='red',size=0.25,linetype='dashed')+
    geom_hline(yintercept = 1.97,color='red',size=0.25,linetype='dashed')+
    geom_hline(yintercept = 0,color='black',size=0.25,linetype='dashed')+
    geom_point(size=3.5)+
    geom_text(aes(label=ID),position = position_identity(),vjust = -1,size=3,color='black')+
    #geom_line(group=1,color='gray44',size=0.15)+
    theme_clean()+
    xlab('')+
    ylab('Fold changes 24m / BL (log2 scale)')+
    scale_color_manual(values=colors_all)+
    scale_fill_manual(values=colors_all)+
    theme(axis.text.y = element_text( size = 10,face = 'bold' ),
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.95, size = 8,face = 'bold'),
          axis.title.x = element_text( size = 8,face = 'bold' ),
          axis.title.y = element_text( size = 8,face = 'bold' ),
          strip.text = element_text(size = 8,face='bold',lineheight=1),
          legend.position = "none",aspect.ratio = 0.25)

o1b <- ggplot(foldChanges_part2) + aes(x=reorder(CurrentState,-LogV4m24),y=LogV4m24,color=Celltype,fill=Celltype,label=ID)+
   geom_hline(yintercept = -2,color='red',size=0.25,linetype='dashed')+
    geom_hline(yintercept = 2,color='red',size=0.25,linetype='dashed')+
    geom_hline(yintercept = 0,color='black',size=0.25,linetype='dashed')+
    geom_point(size=4.5)+
    geom_text(aes(label=ID),position = position_identity(),vjust = -1,size=3,color='black')+
    #geom_line(group=1,color='gray44',size=0.15)+
    theme_clean()+
    xlab('')+
    ylab('Fold changes 24m / BL (log2 scale)')+
    scale_color_manual(values=colors_all)+
    scale_fill_manual(values=colors_all)+
    theme(axis.text.y = element_text( size = 12,face = 'bold' ),
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.95, size = 12,face = 'bold'),
          axis.title.x = element_text( size = 12,face = 'bold' ),
          axis.title.y = element_text( size = 12,face = 'bold' ),
          strip.text = element_text(size = 12,face='bold',lineheight=1),
          legend.position = "none",aspect.ratio = 0.25)

myfile <- paste('plots/Supp_Figure_18.pdf',sep ='')
pdf(myfile,width = 20,height = 7.5)
print(o1b)
dev.off()

o1b <- ggplot(foldChanges_part2) + aes(x=reorder(CurrentState,-LogV4HC),y=LogV4HC,color=Celltype,fill=Celltype,label=ID)+
   geom_hline(yintercept = -2,color='red',size=0.25,linetype='dashed')+
    geom_hline(yintercept = 2,color='red',size=0.25,linetype='dashed')+
    geom_hline(yintercept = 0,color='black',size=0.25,linetype='dashed')+
    geom_point(size=4.5)+
    geom_text(aes(label=ID),position = position_identity(),vjust = -1,size=3,color='black')+
    #geom_line(group=1,color='gray44',size=0.15)+
    theme_clean()+
    xlab('')+
    ylab('Fold changes HC / BL (log2 scale)')+
    scale_color_manual(values=colors_all)+
    scale_fill_manual(values=colors_all)+
    theme(axis.text.y = element_text( size = 12,face = 'bold' ),
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.95, size = 12,face = 'bold'),
          axis.title.x = element_text( size = 12,face = 'bold' ),
          axis.title.y = element_text( size = 12,face = 'bold' ),
          strip.text = element_text(size = 12,face='bold',lineheight=1),
          legend.position = "none",aspect.ratio = 0.25)


#same as in the discovery cohort estimate all differences in the validation cohort
dt <- c(foldChanges_part2$LogV4m24,foldChanges_part2$LogV4HC)
quantile(dt,probs = seq(0,1,0.05))

#threshold for the lower values for the validation cohort, but in the paper we use the cutoffs from the discovery cohort since more differences were computed and thus might have better statistical power
c3 <- quantile(dt,probs = seq(0,1,0.05))[3]
#threshold for the higher values
c4 <- quantile(dt,probs = seq(0,1,0.05))[19]

#visualise all fold changes
colnames(foldChanges_part2)
colnames(foldChanges)[4] <- '24m/BL'
colnames(foldChanges)[5] <- 'HC/BL'

plotData <- data.frame(cellType=foldChanges_part2$Celltype,
                       currentState=foldChanges_part2$CurrentState,
                       BL=0,
                       `24m`= foldChanges_part2$LogV4m24,
                       `HC`= foldChanges_part2$LogV4HC)

promptStates <- filterStates$CurrentState
a <- which(plotData$currentState %in% promptStates)
plotData <- plotData[a,]

colnames(numberOfCells_part2)[1] <- 'currentState'
tmp <- merge(plotData,numberOfCells_part2,by='currentState')


tmp <- melt(plotData,id.vars = c('cellType','currentState'))
tmp$cellType <- factor(tmp$cellType,levels = c('CD4_Memory','CD4_Naive',
                                               'CD8_Memory','CD8_Naive',
                                               'T_DN','T_other',
                                               'B_Memory','B_Naive'))
tmp$currentState <- factor(tmp$currentState,levels=c('CD4_Memory_s7',                                           'CD4_Naive_s2','CD4_Naive_s3','CD4_Naive_s4','CD4_Naive_s6','CD4_Naive_s9',                                                  'CD8_Memory_s2','CD8_Memory_s6','CD8_Naive_s2','CD8_Naive_s3','CD8_Naive_s4','CD8_Naive_s6','T_DN_s3','T_other_s4','B_Memory_s6','B_Memory_s8','B_Naive_s1'))

o1 <- ggplot(tmp) + aes(x=variable,y=value,fill=cellType,color=cellType)+
  geom_hline(yintercept = -1.25,color='red',size=0.15,linetype='dashed')+
  geom_hline(yintercept = 1.97,color='red',size=0.15,linetype='dashed')+
  geom_point(size=3,shape=15)+
  geom_text(aes(label=round(value,2)),position = position_identity(),vjust = -1,size=3,color='black')+
  facet_wrap(~currentState,ncol=6,scales = 'fixed')+
  theme_clean() +
  xlab('')+
  ylim(-8,8)+
  ylab('Validation cohort - Relative changes from BL (log2 scale)')+
  scale_fill_manual(values = colors_all[1:8])+
  scale_color_manual(values = colors_all[1:8])+
  theme(axis.text.y = element_text( size = 8 ),
        axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 0.5, size = 8),
        axis.title.x = element_blank(),
        axis.title.y = element_text( size = 8 ),
        strip.text = element_text(size = 8,face='bold',lineheight=1),
        legend.position = "none",aspect.ratio = 0.8)

myfile <- paste('plots/Figure_5B.pdf',sep ='')
  pdf(myfile,width = 10,height = 7)
  print(o1)
  dev.off()
  
```

