---
title: "High dimensional immune profiling following aHSCT in MS - PART3"
author: "Dimitrios Kleftogiannis"
date: "2024-9-28"
output: html_document
---

### Utility

The data interpretation is performed in collaboration with **Dr. Sonia Gavasso and Jonas Haugsøen**. 

The utility of this code is to reproduce the FreeViz/SARA analysis presented in the paper. 

### Contact

Comments and bug reports are welcome, please email: Dimitrios Kleftogiannis (dimitrios.kleftogiannis@uib.no)

We are also interested to know about how you have used our source code, including any improvements that you have implemented.
 
You are free to modify, extend or distribute our source code, as long as our copyright notice remains unchanged and included in its entirety. 

### License

This code is licensed under the MIT License.

Copyright 2024, University of Bergen (UiB) and Neuro-SysMed, Norway


### Loading packages 
First we load the packages required for the analysis.

```{r load packages, echo=FALSE, eval=FALSE, error=TRUE, warning=FALSE,cache=TRUE}
# set global chunk options
# load libraries
library(RColorBrewer)
library(gplots) 
library(ggplot2)
library(gdata)
library(gridExtra)
library(grid)
library(reshape2)
library(ggthemes)
library(ggpubr)
library(dplyr)
library(Radviz)
library(MASS)
```

### Initialise working directory and load in-house functions
```{r initialise workspace,cache=TRUE,echo=FALSE}
setwd("/Home/siv32/dkl081/Desktop/RAMMS/code_repo")
```

### Specify the markers of interest in the antibody panel
```{r specify marker names,cache=TRUE,echo=FALSE}
#all markers
col.names <- c("CD16","CD4","CD14","CD19","CD3","CD235ab","CD11c","CD33",
               "CD133","CD123","CD162","CD185","CD45RA","CD278","CD194","CD161",
               "CD184","CD27","CD44","CD127","CD10","CD73","HLADR","CD146",
               "CD117","CD8a","CD34","CD105","CD49d","CD20","CD25","CD66b",
               "CD49f","CD45RO","CD90","CD45","CD195","CD38","CD196","CD135","CD56")
col.names


col.names.info <- c('type','type','type','type','type','state','type','type',
                    'state','type','state','state','type','state','state','state',
                    'state','state','state','state','state','state','type','state',
                    'state','type','type','state','state','type','state','type',
                    'state','type','state','type','state','type','state','state','type')

#define the marker_info
marker_info <- data.frame(Channel=NA,
                          marker_name=col.names,
                          marker_class=col.names.info)

```

### RadViz/FreeViz visualisations for selected populations

```{r RadViz/FreeViz analysis, echo=TRUE, cache=TRUE,eval=FALSE}

#name the states you want to analyse
promptStates2 <- c("CD4_Naive_s2",'CD8_Naive_s4','B_Naive_s1',
                   'CD4_Memory_s7','CD8_Memory_s2','B_Memory_s6')   

#load the cells from the validation cohort
load('allLymphocytes_annotated.RDa')
#take out the rest of cells from the discovery cohort
allCells <- allCells[allCells$Sample=='24m'| allCells$Sample=='HC', ]
colnames(allCells)[23] <- 'HLADR'

for(myState in promptStates2){
    
    gc()
    str <- paste('Processing state: ',myState,sep='')
    cat(str)
    cat('')
    #split the name and find the number
    myStateIdx <- stringr::str_extract(myState, "(\\d+$)")
    myStateIdx <- as.integer(myStateIdx)
    myCelltype <- substr(myState, 1, nchar(myState)-3)
    
    myfile <- paste('FlowSOM_allData_states_',myCelltype,'.RDa',sep ='')
    load(myfile) 
    
    myMarkers <- colnames(data_df_test)[1:28]
    myStateData_discovery <- data_df_test[data_df_test$pop_labels==myStateIdx,]      
    #prepare the data for modelling using the lda.model
    myCellTypeData <- allCells[allCells$CellType==myCelltype, ]
    A <- predict(lda.model1,myCellTypeData[,myMarkers])
    myCellTypeData$pop_labels <- A$class
    posteriors <- as.matrix(apply(A$posterior,1,max))
    plot(density(posteriors))
    #filter the data based on prob
    idx1 <- which(posteriors>0.6)
    myCellTypeData <- myCellTypeData[idx1,]
    
    myStateData_validation <- myCellTypeData[myCellTypeData$pop_labels==myStateIdx,]   
    myStateData_validation <- myStateData_validation[,colnames(myStateData_discovery)]
  
    dt <- rbind(myStateData_discovery,myStateData_validation)
    a <- which(dt$Sample=='HC')
    dt <- dt[-a,]
    n <- min(table(dt$Sample))
    #we use 5000 cells if available...otherwise the minimum find
    if(n >= 5000){
      NCells <- 5000
    }else{
      NCells <- n
    }
    
    ###############################################################################
    combinedData <- data.frame()
    combinedDist <- data.frame()
    N <- 50
    for(iter in 1:N){
       
        if(iter%%10==0){
          str <- paste('Iter: ',iter,sep='')
          cat(str)
          gc()
        }
        myStateData.sub <- dt %>% 
            group_by(Sample) %>% 
            sample_n(NCells,replace = FALSE)      
    
        treat.S <- do.optimFreeviz(myStateData.sub[,1:28],classes = myStateData.sub$Sample) 
        data.fv <- do.radviz(myStateData.sub[,1:28], treat.S)   
        data.fv$proj$data$Sample <- factor(myStateData.sub$Sample,levels = c('BL','100d','6m','12m','24m'))
        data.fv <- rescalePlot(data.fv)
        #fetch the data from the FreeViz object
        tmp_dt <- data.frame(data.fv$proj$data)
        combinedData <- rbind(combinedData,tmp_dt)
        tmp_springs <- data.frame(State=myState,
                              Name=rownames(data.fv$springs),
                              X=data.fv$springs[,1],
                              Y=data.fv$springs[,2])
        tmp <- data.fv$springs
        #compute euclidean distance that quantifies the contribution of the marker 
        d <- matrix(-1,nrow = nrow(tmp),ncol = 1)
        for(i in 1:nrow(tmp)){
            d[i,1] <- sqrt(sum((tmp[i,1] - tmp[i,2])^2))
        }
        tmp_springs$Dist <- d[,1]
        if(iter==1){
          combinedDist <- tmp_springs$Dist
        }else{
          combinedDist <- cbind(combinedDist,tmp_springs$Dist)
        }
    }
    #work with the distances
    rownames(combinedDist) <- tmp_springs$Name
    #generate the average distance
    combinedDist <- data.frame(combinedDist)
    combinedDist$Name <- rownames(combinedDist)
    avgDist <- rowMeans(combinedDist[,1:N])
    SD <- apply(combinedDist[,1:N],1, sd, na.rm = TRUE)
    SE <- SD/sqrt(N)
    combinedDist$Avg <- avgDist
    combinedDist$SE <- SE
    
    o3 <- ggplot(combinedDist) + aes(x=reorder(Name,-Avg),y=Avg)+
              geom_point(size=1.5,color='royalblue4')+
              geom_errorbar( aes(ymin = Avg-SE, ymax = Avg+SE),width = 0.3) +
              theme_clean()+
              ggtitle(myState)+
              xlab('')+
              ylab('Average euclidean distance with SE')+
              theme(axis.text.y = element_text( size = 12,face = 'bold' ),
                    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.95, size = 12,face = 'bold'),
                    axis.title.x = element_text( size = 12,face = 'bold' ),
                    axis.title.y = element_text( size = 12,face = 'bold' ),
                    strip.text = element_text(size = 12,face='bold',lineheight=1),
                    legend.position = "none",aspect.ratio = 0.25)
    
    ###############################################################################
    combinedData$Sample <- factor(combinedData$Sample,levels=c('BL','100d','6m','12m','24m'))
    
    o2 <- ggplot(combinedData, aes(rx,ry,color=Sample))+
            #geom_point()+              
            geom_density_2d(,linewidth=0.8)+
                    #facet_wrap(~Sample,nrow=1)+
                    geom_hline(yintercept = 0,color='black',size=0.15,linetype='dashed')+
                    geom_vline(xintercept = 0,color='black',size=0.15,linetype='dashed')+
                    ggtitle(myState)+
                    theme_radviz(base_size = 16)+
                    theme(axis.text.y = element_blank(),
                          axis.text.x = element_blank(),
                          axis.title.x = element_blank(),
                          axis.title.y = element_blank(),
                          legend.position = "bottom",
                          aspect.ratio = 1)+
                    scale_color_manual(values = c("deepskyblue2","magenta2","indianred2","rosybrown2", "rosybrown4"))+
                    guides(color = guide_legend(override.aes = list(size=18),nrow=1,title=""))
    
      ###############################################################################
      # below we save the generated plots and produce marker expression plots for selected markers
      ###############################################################################
    
    if(myState=='CD4_Naive_s2'){
      
      myfile <- paste('plots/Figure_6a.pdf',sep ='')
      pdf(myfile)
      print(o2)
      dev.off()
      
       myfile <- paste('plots/Supp_Figure_19a.pdf',sep ='')
       pdf(myfile)
       print(o3)
       dev.off()
 
      tmp <- dt[,c('CD194','CD38','CD235ab','CD127','Sample')]
      tmp <- melt(tmp)
      tmp$Sample <- factor( tmp$Sample,levels = c('BL','100d','6m','12m','24m'))
      o <- ggplot(tmp,aes(x=Sample,y=value,fill=Sample,color=Sample))+
              geom_boxplot(width=0.15,alpha=0.8,size=0.45,color='black',
                 outlier.colour = "black",
                 outlier.shape =NA,
                 outlier.fill = "red",
                 outlier.size = 0.05,notch = FALSE)+
                 facet_wrap(~variable,nrow=1,scales ='free_x')+
              scale_color_manual(values = c("deepskyblue2","magenta2","indianred2","rosybrown2", "rosybrown4"))+
              ylab('Expression')+
              ggtitle(myState)+
              theme_clean() +
              ylim(0,0.4)+
              theme(axis.text.y = element_text( size = 12 ),
                    axis.text.x = element_text(angle = 0, vjust = 1, hjust = 0.5, size = 12),
                    axis.title.x = element_blank(),
                    axis.title.y = element_text( size = 12 ),
                    strip.text = element_text(size = 12,face='bold',lineheight=1),
                    legend.position = "none",aspect.ratio = 0.81)
      
      myfile <- paste('plots/Supp_Figure_20a.pdf',sep ='')
      pdf(myfile,width = 9)
      print(o)
      dev.off()
    }
    
    if(myState=='CD8_Naive_s4'){
      
        myfile <- paste('plots/Figure_6b.pdf',sep ='')
        pdf(myfile)
        print(o2)
        dev.off()
      
        myfile <- paste('plots/Supp_Figure_19b.pdf',sep ='')
        pdf(myfile)
        print(o3)
        dev.off()
      
      tmp <- dt[,c('CD27','CD194','CD195','CD49d','Sample')]
      tmp <- melt(tmp)
      tmp$Sample <- factor( tmp$Sample,levels = c('BL','100d','6m','12m','24m'))
      o <- ggplot(tmp,aes(x=Sample,y=value,fill=Sample,color=Sample))+
              geom_boxplot(width=0.15,alpha=0.8,size=0.45,color='black',
                 outlier.colour = "black",
                 outlier.shape =NA,
                 outlier.fill = "red",
                 outlier.size = 0.05,notch = FALSE)+
                 facet_wrap(~variable,nrow=1,scales ='free_x')+
              scale_color_manual(values = c("deepskyblue2","magenta2","indianred2","rosybrown2", "rosybrown4"))+
              ylab('Expression')+
              ggtitle(myState)+
              theme_clean() +
              ylim(0,0.8)+
              theme(axis.text.y = element_text( size = 12 ),
                    axis.text.x = element_text(angle = 0, vjust = 1, hjust = 0.5, size = 12),
                    axis.title.x = element_blank(),
                    axis.title.y = element_text( size = 12 ),
                    strip.text = element_text(size = 12,face='bold',lineheight=1),
                    legend.position = "none",aspect.ratio = 0.81)
      
      myfile <- paste('plots/Supp_Figure_20b.pdf',sep ='')
      pdf(myfile,width = 9)
      print(o)
      dev.off()
    }
      
    if(myState=='B_Naive_s1'){
      
        myfile <- paste('plots/Figure_6c.pdf',sep ='')
        pdf(myfile)
        print(o2)
        dev.off()
      
        myfile <- paste('plots/Supp_Figure_19c.pdf',sep ='')
        pdf(myfile)
        print(o3)
        dev.off()
      
      tmp <- dt[,c('CD38','CD235ab','CD196','HLADR','Sample')]
      tmp <- melt(tmp)
      tmp$Sample <- factor( tmp$Sample,levels = c('BL','100d','6m','12m','24m'))
      o <- ggplot(tmp,aes(x=Sample,y=value,fill=Sample,color=Sample))+
              geom_boxplot(width=0.15,alpha=0.8,size=0.45,color='black',
                 outlier.colour = "black",
                 outlier.shape =NA,
                 outlier.fill = "red",
                 outlier.size = 0.05,notch = FALSE)+
                 facet_wrap(~variable,nrow=1,scales ='free_x')+
              scale_color_manual(values = c("deepskyblue2","magenta2","indianred2","rosybrown2", "rosybrown4"))+
              ylab('Expression')+
              ggtitle(myState)+
              theme_clean() +
              ylim(0,0.8)+
              theme(axis.text.y = element_text( size = 12 ),
                    axis.text.x = element_text(angle = 0, vjust = 1, hjust = 0.5, size = 12),
                    axis.title.x = element_blank(),
                    axis.title.y = element_text( size = 12 ),
                    strip.text = element_text(size = 12,face='bold',lineheight=1),
                    legend.position = "none",aspect.ratio = 0.81)
      
      myfile <- paste('plots/Supp_Figure_20c.pdf',sep ='')
      pdf(myfile,width = 9)
      print(o)
      dev.off()
    }
   
     if(myState=='CD4_Memory_s7'){
      
       myfile <- paste('plots/Figure_6d.pdf',sep ='')
       pdf(myfile)
       print(o2)
       dev.off()
      
       myfile <- paste('plots/Supp_Figure_19d.pdf',sep ='')
       pdf(myfile)
       print(o3)
       dev.off()
       
      tmp <- dt[,c('CD27','CD194','CD161','CD195','Sample')]
      tmp <- melt(tmp)
      tmp$Sample <- factor( tmp$Sample,levels = c('BL','100d','6m','12m','24m'))
     o <- ggplot(tmp,aes(x=Sample,y=value,fill=Sample,color=Sample))+
              geom_boxplot(width=0.15,alpha=0.8,size=0.45,color='black',
                 outlier.colour = "black",
                 outlier.shape =NA,
                 outlier.fill = "red",
                 outlier.size = 0.05,notch = FALSE)+
                 facet_wrap(~variable,nrow=1,scales ='free_x')+
              scale_color_manual(values = c("deepskyblue2","magenta2","indianred2","rosybrown2", "rosybrown4"))+
              ylab('Expression')+
              ggtitle(myState)+
              theme_clean() +
              ylim(0,0.9)+
              theme(axis.text.y = element_text( size = 12 ),
                    axis.text.x = element_text(angle = 0, vjust = 1, hjust = 0.5, size = 12),
                    axis.title.x = element_blank(),
                    axis.title.y = element_text( size = 12 ),
                    strip.text = element_text(size = 12,face='bold',lineheight=1),
                    legend.position = "none",aspect.ratio = 0.81)
      
        myfile <- paste('plots/Supp_Figure_20d.pdf',sep ='')
        pdf(myfile,width = 9)
        print(o)
        dev.off()
     
    }
   
 
    if(myState=='CD8_Memory_s2'){
      
        myfile <- paste('plots/Figure_6e.pdf',sep ='')
        pdf(myfile)
        print(o2)
        dev.off()
      
        myfile <- paste('plots/Supp_Figure_19e.pdf',sep ='')
        pdf(myfile)
        print(o3)
        dev.off()
      
      tmp <- dt[,c('CD27','CD127','CD49d','CD195','Sample')]
      tmp <- melt(tmp)
      tmp$Sample <- factor( tmp$Sample,levels = c('BL','100d','6m','12m','24m'))
      o <- ggplot(tmp,aes(x=Sample,y=value,fill=Sample,color=Sample))+
              geom_boxplot(width=0.15,alpha=0.8,size=0.45,color='black',
                 outlier.colour = "black",
                 outlier.shape =NA,
                 outlier.fill = "red",
                 outlier.size = 0.05,notch = FALSE)+
                 facet_wrap(~variable,nrow=1,scales ='free_x')+
              scale_color_manual(values = c("deepskyblue2","magenta2","indianred2","rosybrown2", "rosybrown4"))+
              ylab('Expression')+
              ggtitle(myState)+
              theme_clean() +
              ylim(0,0.9)+
              theme(axis.text.y = element_text( size = 12 ),
                    axis.text.x = element_text(angle = 0, vjust = 1, hjust = 0.5, size = 12),
                    axis.title.x = element_blank(),
                    axis.title.y = element_text( size = 12 ),
                    strip.text = element_text(size = 12,face='bold',lineheight=1),
                    legend.position = "none",aspect.ratio = 0.81)
        
        myfile <- paste('plots/Supp_Figure_20e.pdf',sep ='')
        pdf(myfile,width = 9)
        print(o)
        dev.off()
      
    }
    
    
    if(myState=='B_Memory_s6'){
      
       myfile <- paste('plots/Figure_6f.pdf',sep ='')
      pdf(myfile)
      print(o2)
      dev.off()
      
       myfile <- paste('plots/Supp_Figure_19f.pdf',sep ='')
       pdf(myfile)
       print(o3)
       dev.off()
      
      tmp <- dt[,c('CD194','CD38','HLADR','CD44','Sample')]
      tmp <- melt(tmp)
      tmp$Sample <- factor( tmp$Sample,levels = c('BL','100d','6m','12m','24m'))
     o <- ggplot(tmp,aes(x=Sample,y=value,fill=Sample,color=Sample))+
              geom_boxplot(width=0.15,alpha=0.8,size=0.45,color='black',
                 outlier.colour = "black",
                 outlier.shape =NA,
                 outlier.fill = "red",
                 outlier.size = 0.05,notch = FALSE)+
                 facet_wrap(~variable,nrow=1,scales ='free_x')+
              scale_color_manual(values = c("deepskyblue2","magenta2","indianred2","rosybrown2", "rosybrown4"))+
              ylab('Expression')+
              ggtitle(myState)+
              theme_clean() +
              ylim(0,0.9)+
              theme(axis.text.y = element_text( size = 12 ),
                    axis.text.x = element_text(angle = 0, vjust = 1, hjust = 0.5, size = 12),
                    axis.title.x = element_blank(),
                    axis.title.y = element_text( size = 12 ),
                    strip.text = element_text(size = 12,face='bold',lineheight=1),
                    legend.position = "none",aspect.ratio = 0.81)
      
    
        myfile <- paste('plots/Supp_Figure_20f.pdf',sep ='')
        pdf(myfile,width = 9)
        print(o)
        dev.off()
    }
}
```



### SARA analysis for previous populations and top-ranked markers
The SARA code was adopted from its original implementation in MATLAB. For simplicity we compare BL vs. 24m to quantify changes before aHSCT and long after it. 

```{r SARA, echo=TRUE, cache=TRUE,eval=FALSE}

source('utilityFunc.R')
promptStates2 <- c("CD4_Naive_s2",'CD8_Naive_s4','B_Naive_s1',
                   'CD4_Memory_s7','CD8_Memory_s2','B_Memory_s6')  

#number of permutations 
N <- 10000
br <- data.frame()
for(myState in promptStates2){
    
    str <- paste('Processing state: ',myState,sep='')
    cat(str)
    cat('')
  
    myStateIdx <- stringr::str_extract(myState, "(\\d+$)")
    myStateIdx <- as.integer(myStateIdx)
    myCelltype <- substr(myState, 1, nchar(myState)-3)
    
    myfile <- paste('FlowSOM_allData_states_',myCelltype,'.RDa',sep ='')
    load(myfile) 
    
    myMarkers <- colnames(data_df_test)[1:28]
    myStateData_discovery <- data_df_test[data_df_test$pop_labels==myStateIdx,]      
    #prepare the data for modelling using the lda.model
    myCellTypeData <- allCells[allCells$CellType==myCelltype, ]
    A <- predict(lda.model1,myCellTypeData[,myMarkers])
    myCellTypeData$pop_labels <- A$class
    posteriors <- as.matrix(apply(A$posterior,1,max))
    plot(density(posteriors))
    #filter the data based on prob
    idx1 <- which(posteriors>0.6)
    myCellTypeData <- myCellTypeData[idx1,]
    myStateData_validation <- myCellTypeData[myCellTypeData$pop_labels==myStateIdx,]   
    myStateData_validation <- myStateData_validation[,colnames(myStateData_discovery)]
    dt <- rbind(myStateData_discovery,myStateData_validation)
    a <- which(dt$Sample=='HC')
    dt <- dt[-a,]
    n <- min(table(dt$Sample))
    #select equal amount of cells
    if(n >= 10000){
      NCells <- 10000
    }else{
      NCells <- n
    }
    
    NCells 
    dt <- dt %>% 
        group_by(Sample) %>% 
        sample_n(NCells,replace = FALSE)     
    dt <- as.data.frame(dt)
    
    if(myState=='CD4_Naive_s2'){
      summaryAllSARA <- data.frame()
      #stateMarkerNames <- c('CD194','CD38','CD235ab','CD127')
      stateMarkerNames <- myMarkers
      IDX <- 1
      SARAscores <- matrix(-1,nrow = 1,ncol=length(stateMarkerNames))
      SARApvalues <- matrix(-1,nrow = 1,ncol=length(stateMarkerNames))
      for(myMarker in stateMarkerNames){
      
        str <- paste('  Processing marker:',myMarker,sep='')
        cat(str)
        cat('\n') 
     
        channelA <- dt[dt$Sample=='BL',myMarker]
        channelB <- dt[dt$Sample=='24m',myMarker]
        #generate the ecdf
        ecdfA<- ecdf(channelA)
        ecdfB<- ecdf(channelB)
        A <- ecdfA(channelA)
        B <- ecdfB(channelB)
        myEMD <- suppressWarnings(manhattan_dist(A,B))
        #perform permutations
        myCounter <- 0
        for(myPerm in 1:N){
      
            #shuffle the condition channel and generate random ecdf
            C <- sample(channelB)
            ecdfC <- ecdf(C)
            myC <- ecdfC(C)
            #permuted 
            permEMD <- suppressWarnings(manhattan_dist(A,myC))
            if(permEMD>myEMD){
                  myCounter <- myCounter + 1
            }
        }
        pvalue <- myCounter/N
        pvalue
        #myEMD is a distance --> larger distance more different the samples
        #sign(mean(channelB)-mean(channelA) --> +1 if mean at the condition is greater than mean at baseline
        #1-pvalue is a score --> small p value means more significant changes so 1-p as score gets higher
        #overall higher SARA score means greater differences between the two experimental conditions
        myScore <- myEMD*sign(mean(channelB,na.rm = T)-mean(channelA,na.rm = T))*(1-pvalue)
        SARAscores[1,IDX] <- myScore
        SARApvalues[1,IDX] <- pvalue
        IDX <- IDX + 1
    }
    SARAscores <- data.frame(SARAscores)
    colnames(SARAscores) <- stateMarkerNames
    SARAscores$CellType <- myState
    #SARAscores$pvalue <- pvalue
    summaryAllSARA <- rbind(summaryAllSARA,SARAscores)
    gc()
    br <- rbind(br,melt(summaryAllSARA))
    }
    ###########################################################
    if(myState=='CD8_Naive_s4'){
      summaryAllSARA <- data.frame()
      #stateMarkerNames <- c('CD27','CD194','CD49d','CD195')
       stateMarkerNames <- myMarkers
      IDX <- 1
      SARAscores <- matrix(-1,nrow = 1,ncol=length(stateMarkerNames))
      SARApvalues <- matrix(-1,nrow = 1,ncol=length(stateMarkerNames))
      for(myMarker in stateMarkerNames){
      
        str <- paste('  Processing marker:',myMarker,sep='')
        cat(str)
        cat('\n') 
     
        channelA <- dt[dt$Sample=='BL',myMarker]
        channelB <- dt[dt$Sample=='24m',myMarker]
        #generate the ecdf
        ecdfA<- ecdf(channelA)
        ecdfB<- ecdf(channelB)
        A <- ecdfA(channelA)
        B <- ecdfB(channelB)
        myEMD <- suppressWarnings(manhattan_dist(A,B))
        #perform permutations
        myCounter <- 0
        for(myPerm in 1:N){
      
            #shuffle the condition channel and generate random ecdf
            C <- sample(channelB)
            ecdfC <- ecdf(C)
            myC <- ecdfC(C)
            #permuted 
            permEMD <- suppressWarnings(manhattan_dist(A,myC))
            if(permEMD>myEMD){
                  myCounter <- myCounter + 1
            }
        }
        pvalue <- myCounter/N
        pvalue
        #myEMD is a distance --> larger distance more different the samples
        #sign(mean(channelB)-mean(channelA) --> +1 if mean at the condition is greater than mean at baseline
        #1-pvalue is a score --> small p value means more significant changes so 1-p as score gets higher
        #overall higher SARA score means greater differences between the two experimental conditions
        myScore <- myEMD*sign(mean(channelB,na.rm = T)-mean(channelA,na.rm = T))*(1-pvalue)
        SARAscores[1,IDX] <- myScore
        SARApvalues[1,IDX] <- pvalue
        IDX <- IDX + 1
    }
    SARAscores <- data.frame(SARAscores)
    colnames(SARAscores) <- stateMarkerNames
    SARAscores$CellType <- myState
    #SARAscores$pvalue <- pvalue
    summaryAllSARA <- rbind(summaryAllSARA,SARAscores)
    br <- rbind(br,melt(summaryAllSARA))
    gc()
    }
    ###########################################################
    if(myState=='B_Naive_s1'){
      summaryAllSARA <- data.frame()
      #stateMarkerNames <- c('CD196','CD38','HLADR','CD235ab')
      stateMarkerNames <- myMarkers
      IDX <- 1
      SARAscores <- matrix(-1,nrow = 1,ncol=length(stateMarkerNames))
      SARApvalues <- matrix(-1,nrow = 1,ncol=length(stateMarkerNames))
      for(myMarker in stateMarkerNames){
      
        str <- paste('  Processing marker:',myMarker,sep='')
        cat(str)
        cat('\n') 
     
        channelA <- dt[dt$Sample=='BL',myMarker]
        channelB <- dt[dt$Sample=='24m',myMarker]
        #generate the ecdf
        ecdfA<- ecdf(channelA)
        ecdfB<- ecdf(channelB)
        A <- ecdfA(channelA)
        B <- ecdfB(channelB)
        myEMD <- suppressWarnings(manhattan_dist(A,B))
        #perform permutations
        myCounter <- 0
        for(myPerm in 1:N){
      
            #shuffle the condition channel and generate random ecdf
            C <- sample(channelB)
            ecdfC <- ecdf(C)
            myC <- ecdfC(C)
            #permuted 
            permEMD <- suppressWarnings(manhattan_dist(A,myC))
            if(permEMD>myEMD){
                  myCounter <- myCounter + 1
            }
        }
        pvalue <- myCounter/N
        pvalue
        #myEMD is a distance --> larger distance more different the samples
        #sign(mean(channelB)-mean(channelA) --> +1 if mean at the condition is greater than mean at baseline
        #1-pvalue is a score --> small p value means more significant changes so 1-p as score gets higher
        #overall higher SARA score means greater differences between the two experimental conditions
        myScore <- myEMD*sign(mean(channelB,na.rm = T)-mean(channelA,na.rm = T))*(1-pvalue)
        SARAscores[1,IDX] <- myScore
        SARApvalues[1,IDX] <- pvalue
        IDX <- IDX + 1
    }
    SARAscores <- data.frame(SARAscores)
    colnames(SARAscores) <- stateMarkerNames
    SARAscores$CellType <- myState
    #SARAscores$pvalue <- pvalue
    summaryAllSARA <- rbind(summaryAllSARA,SARAscores)
    gc()
     br <- rbind(br,melt(summaryAllSARA))
    }
    ###########################################################
     if(myState=='CD4_Memory_s7'){
      summaryAllSARA <- data.frame()
      #stateMarkerNames <- c('CD27','CD195','CD161','CD194')
      stateMarkerNames <- myMarkers
      IDX <- 1
      SARAscores <- matrix(-1,nrow = 1,ncol=length(stateMarkerNames))
      SARApvalues <- matrix(-1,nrow = 1,ncol=length(stateMarkerNames))
      for(myMarker in stateMarkerNames){
      
        str <- paste('  Processing marker:',myMarker,sep='')
        cat(str)
        cat('\n') 
     
        channelA <- dt[dt$Sample=='BL',myMarker]
        channelB <- dt[dt$Sample=='24m',myMarker]
        #generate the ecdf
        ecdfA<- ecdf(channelA)
        ecdfB<- ecdf(channelB)
        A <- ecdfA(channelA)
        B <- ecdfB(channelB)
        myEMD <- suppressWarnings(manhattan_dist(A,B))
        #perform permutations
        myCounter <- 0
        for(myPerm in 1:N){
      
            #shuffle the condition channel and generate random ecdf
            C <- sample(channelB)
            ecdfC <- ecdf(C)
            myC <- ecdfC(C)
            #permuted 
            permEMD <- suppressWarnings(manhattan_dist(A,myC))
            if(permEMD>myEMD){
                  myCounter <- myCounter + 1
            }
        }
        pvalue <- myCounter/N
        pvalue
        #myEMD is a distance --> larger distance more different the samples
        #sign(mean(channelB)-mean(channelA) --> +1 if mean at the condition is greater than mean at baseline
        #1-pvalue is a score --> small p value means more significant changes so 1-p as score gets higher
        #overall higher SARA score means greater differences between the two experimental conditions
        myScore <- myEMD*sign(mean(channelB,na.rm = T)-mean(channelA,na.rm = T))*(1-pvalue)
        SARAscores[1,IDX] <- myScore
        SARApvalues[1,IDX] <- pvalue
        IDX <- IDX + 1
    }
    SARAscores <- data.frame(SARAscores)
    colnames(SARAscores) <- stateMarkerNames
    SARAscores$CellType <- myState
    #SARAscores$pvalue <- pvalue
    summaryAllSARA <- rbind(summaryAllSARA,SARAscores)
    gc()
     br <- rbind(br,melt(summaryAllSARA))
    }
    ###########################################################
    if(myState=='CD8_Memory_s2'){
      summaryAllSARA <- data.frame()
      #stateMarkerNames <- c('CD127','CD27','CD195','CD49d')
      stateMarkerNames <- myMarkers
      IDX <- 1
      SARAscores <- matrix(-1,nrow = 1,ncol=length(stateMarkerNames))
      SARApvalues <- matrix(-1,nrow = 1,ncol=length(stateMarkerNames))
      for(myMarker in stateMarkerNames){
      
        str <- paste('  Processing marker:',myMarker,sep='')
        cat(str)
        cat('\n') 
     
        channelA <- dt[dt$Sample=='BL',myMarker]
        channelB <- dt[dt$Sample=='24m',myMarker]
        #generate the ecdf
        ecdfA<- ecdf(channelA)
        ecdfB<- ecdf(channelB)
        A <- ecdfA(channelA)
        B <- ecdfB(channelB)
        myEMD <- suppressWarnings(manhattan_dist(A,B))
        #perform permutations
        myCounter <- 0
        for(myPerm in 1:N){
      
            #shuffle the condition channel and generate random ecdf
            C <- sample(channelB)
            ecdfC <- ecdf(C)
            myC <- ecdfC(C)
            #permuted 
            permEMD <- suppressWarnings(manhattan_dist(A,myC))
            if(permEMD>myEMD){
                  myCounter <- myCounter + 1
            }
        }
        pvalue <- myCounter/N
        pvalue
        #myEMD is a distance --> larger distance more different the samples
        #sign(mean(channelB)-mean(channelA) --> +1 if mean at the condition is greater than mean at baseline
        #1-pvalue is a score --> small p value means more significant changes so 1-p as score gets higher
        #overall higher SARA score means greater differences between the two experimental conditions
        myScore <- myEMD*sign(mean(channelB,na.rm = T)-mean(channelA,na.rm = T))*(1-pvalue)
        SARAscores[1,IDX] <- myScore
        SARApvalues[1,IDX] <- pvalue
        IDX <- IDX + 1
    }
    SARAscores <- data.frame(SARAscores)
    colnames(SARAscores) <- stateMarkerNames
    SARAscores$CellType <- myState
    #SARAscores$pvalue <- pvalue
    summaryAllSARA <- rbind(summaryAllSARA,SARAscores)
    gc()
     br <- rbind(br,melt(summaryAllSARA))
    }
    ###########################################################
     if(myState=='B_Memory_s6'){
      summaryAllSARA <- data.frame()
      #stateMarkerNames <- c('CD194','CD38','HLADR','CD44')
      stateMarkerNames <- myMarkers
      IDX <- 1
      SARAscores <- matrix(-1,nrow = 1,ncol=length(stateMarkerNames))
      SARApvalues <- matrix(-1,nrow = 1,ncol=length(stateMarkerNames))
      for(myMarker in stateMarkerNames){
      
        str <- paste('  Processing marker:',myMarker,sep='')
        cat(str)
        cat('\n') 
     
        channelA <- dt[dt$Sample=='BL',myMarker]
        channelB <- dt[dt$Sample=='24m',myMarker]
        #generate the ecdf
        ecdfA<- ecdf(channelA)
        ecdfB<- ecdf(channelB)
        A <- ecdfA(channelA)
        B <- ecdfB(channelB)
        myEMD <- suppressWarnings(manhattan_dist(A,B))
        #perform permutations
        myCounter <- 0
        for(myPerm in 1:N){
      
            #shuffle the condition channel and generate random ecdf
            C <- sample(channelB)
            ecdfC <- ecdf(C)
            myC <- ecdfC(C)
            #permuted 
            permEMD <- suppressWarnings(manhattan_dist(A,myC))
            if(permEMD>myEMD){
                  myCounter <- myCounter + 1
            }
        }
        pvalue <- myCounter/N
        pvalue
        #myEMD is a distance --> larger distance more different the samples
        #sign(mean(channelB)-mean(channelA) --> +1 if mean at the condition is greater than mean at baseline
        #1-pvalue is a score --> small p value means more significant changes so 1-p as score gets higher
        #overall higher SARA score means greater differences between the two experimental conditions
        myScore <- myEMD*sign(mean(channelB,na.rm = T)-mean(channelA,na.rm = T))*(1-pvalue)
        SARAscores[1,IDX] <- myScore
        SARApvalues[1,IDX] <- pvalue
        IDX <- IDX + 1
    }
    SARAscores <- data.frame(SARAscores)
    colnames(SARAscores) <- stateMarkerNames
    SARAscores$CellType <- myState
    #SARAscores$pvalue <- pvalue
    summaryAllSARA <- rbind(summaryAllSARA,SARAscores)
    gc()
     br <- rbind(br,melt(summaryAllSARA))
    }
    ###########################################################
    
    
    
   

}

br$CellType <- factor(br$CellType,levels=c("CD4_Naive_s2",'CD8_Naive_s4','B_Naive_s1',
                   'CD4_Memory_s7','CD8_Memory_s2','B_Memory_s6'))
 #visualise the SARA scores 
    o <- ggplot(br) + aes(x=variable,y=value,fill=CellType)+
          geom_bar(stat='identity',position='stack',width = 0.48,alpha=1,color='black',size=0.15)+
          facet_wrap(~CellType,scales='free_y',nrow=2)+
          geom_hline(yintercept = 0,color='red',size=0.25)+
          theme_clean()+
      ylab('SARA score')+
      xlab('')+
      theme(axis.text.y = element_text( size = 10 ),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.5, size = 10,face = 'bold'),
        axis.title.x = element_text( size = 10,face = 'bold' ),
        axis.title.y = element_text( size = 12,face='bold' ),
        strip.text = element_text(size = 12,face='bold',lineheight=1),
        legend.position = "none",aspect.ratio = 0.8)+
  #scale_fill_manual(values=myPalete)+
  guides(fill = guide_legend(override.aes = list(size=1.5),nrow=2,title=""))

     myfile <- paste('plots/Supp_Figure_21.pdf',sep ='')
     pdf(myfile,width = 18,height = 10)
     print(o)
     dev.off()

```


