---
title: "High dimensional immune profiling following aHSCT in MS - PART1"
author: "Dimitrios Kleftogiannis"
date: "2024-8-30"
output: html_document
---

### Utility

The utility of this code is to load the processed data and generate visualisations that correspond to Figure 1-3 of the manuscript.

### Contact

Comments and bug reports are welcome, please email: Dimitrios Kleftogiannis (dimitrios.kleftogiannis@uib.no)

We are also interested to know about how you have used our source code, including any improvements that you have implemented.
 
You are free to modify, extend or distribute our source code, as long as our copyright notice remains unchanged and included in its entirety. 

### License

This code is licensed under the MIT License.

Copyright 2024, University of Bergen (UiB) and Neuro-SysMed, Norway


### Loading packages 
First we load the packages required for the analysis.

```{r load packages, echo=FALSE, eval=FALSE, error=TRUE, warning=FALSE,cache=TRUE}
# set global chunk options
# load libraries
library(RColorBrewer)
library(gplots) 
library(ggplot2)
library(gdata)
library(gridExtra)
library(grid)
library(reshape2)
library(ggthemes)
library(ggpubr)
library(dplyr)
library(umap)
library(ComplexHeatmap)
library(ConsensusClusterPlus)
library(readxl)
library(igraph)
library(matrixStats)
library(ggforce)
library(ggraph)
library(ggridges)
```

### Initialise working directory and load in-house functions
```{r initialise workspace,cache=TRUE,echo=FALSE}
setwd("/Home/siv32/dkl081/Desktop/RAMMS/code_repo")
```

### Specify the markers of interest in the antibody panel
```{r specify marker names,cache=TRUE,echo=FALSE}
#all markers
col.names <- c("CD16","CD4","CD14","CD19","CD3","CD235ab","CD11c","CD33",
               "CD133","CD123","CD162","CD185","CD45RA","CD278","CD194","CD161",
               "CD184","CD27","CD44","CD127","CD10","CD73","HLADR","CD146",
               "CD117","CD8a","CD34","CD105","CD49d","CD20","CD25","CD66b",
               "CD49f","CD45RO","CD90","CD45","CD195","CD38","CD196","CD135","CD56")
col.names


col.names.info <- c('type','type','type','type','type','state','type','type',
                    'state','type','state','state','type','state','state','state',
                    'state','state','state','state','state','state','type','state',
                    'state','type','type','state','state','type','state','type',
                    'state','type','state','type','state','type','state','state','type')

#define the marker_info
marker_info <- data.frame(Channel=NA,
                          marker_name=col.names,
                          marker_class=col.names.info)

```


### Generate summary plot showing the total number of cells found in the cohort (all patients and all time points)

```{r generate plots withg absolute number of cells, cache=TRUE,echo=FALSE}

#eliminate scientific notation
options(scipen = 999)

#colors for the time points 
color_TimePoints <- c('gray48',
                      "deepskyblue2",
                      "magenta2",
                      "indianred2",
                      "rosybrown2",
                      "rosybrown4")

TimePointLevels <- c('HC','BL','100d','6m','12m','24m')

#colors for the cell types
colors_all <- c('slategray1','slategray3',
                'steelblue1','steelblue3',
                'darkslategrey','slateblue4',
                'firebrick1','firebrick4','indianred1',
                'orange','orange3','peachpuff',
                'palegreen3','palegreen1','lightgoldenrod3','lightgoldenrod1','greenyellow')

#load the annotated data
load('/Home/siv32/dkl081/Desktop/RAMMS/code_repo/allLymphocytes_annotated.RDa')

#total number of cells in the cohort
timePointCells <- allCells %>% 
  group_by(Sample) %>% 
  summarise(Cells = n())

timePointCells$Sample <- factor(timePointCells$Sample,levels = TimePointLevels)

str <- paste('Number of lymphocytes',sep='')
o1 <- ggplot(timePointCells) + aes(x=Sample,y=Cells,fill=Sample)+
  geom_bar(stat='identity',width = 0.38,alpha=1,color='black',size=0.2)+
  ylab(str)+
  theme_bw()+
  scale_fill_manual(values = color_TimePoints)+
  theme(legend.position = "none",
        axis.title.y = element_text(size = 16),
        axis.text.y = element_text(size = 12),
        axis.text.x = element_text(angle = 0, hjust = 0.5, size = 12),
        axis.title.x = element_blank(),aspect.ratio = 0.5)

myfile <- paste('plots/Number_of_lymphocytes.pdf',sep ='')
pdf(myfile)
print(o1)
dev.off()


#plot the number of cells per cell type
  timePointCellType <- allCells%>% 
  group_by(CellType,Sample) %>% 
  summarise(samples = n())

#fix the order of cell types in the plot
all_cellType_order <- c("CD4_Memory" ,"CD4_Naive" ,
                        "CD8_Memory","CD8_Naive",
                        "T_DN","T_other",
                        "B_Memory" ,"B_Naive" ,'B_other',
                        "NK_CD16_pos", "NK_CD16_neg", "NK_other",
                        "Monocytes_classical", "Monocytes_nonClassical", "DC","DC_plasmocytoid","Myelocytes_other")

timePointCellType$CellType <- factor(timePointCellType$CellType,levels = all_cellType_order)
timePointCellType$Sample <- factor(timePointCellType$Sample,levels = TimePointLevels)

str <- paste('Number of cells',sep='')
o2 <- ggplot(timePointCellType) + aes(x=CellType,y=samples,fill=CellType)+
  geom_bar(stat='identity',width = 0.28,size=0.2,alpha=1,color='black',size=0.2)+
  facet_wrap(~ Sample, scales = "free_y", nrow = 3)+
  ylab(str)+
  theme_bw()+
  scale_fill_manual(values = colors_all)+
  theme(legend.position = "none",
        axis.title.y = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.text.x = element_text(angle = 90, size = 12,hjust = 0.95,vjust = 0.5),
        axis.title.x = element_blank(),aspect.ratio = 0.5)

myfile <- paste('plots/Supp_Figure_7.pdf',sep ='')
pdf(myfile)
print(o2)
dev.off()

```

### Relative abundance of lymphocyte cell populations in the cohort
```{r generate plots withg relative abundance,cache=TRUE,echo=FALSE}

#redo the plot total number of cells per cell population
timePointCellTypeFreq <- allCells %>% 
  group_by(Sample,CellType) %>% 
  summarise(cells = n())%>% 
  mutate(freq = cells/sum(cells))

timePointCellTypeFreq$CellType <- factor(timePointCellTypeFreq$CellType,levels = all_cellType_order)
timePointCellTypeFreq$Sample <- factor(timePointCellTypeFreq$Sample,levels = TimePointLevels)

str <- paste('Relative abundance',sep='')

relative_abundance_all <- ggplot(timePointCellTypeFreq) + aes(x=Sample,y=freq,fill=CellType)+
  geom_bar(stat='identity',width = 0.58,alpha=1,color='black',size=0.2)+
  ylab(str)+
  theme_bw()+
  theme(axis.text.y = element_text( size = 12 ),
        axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 0.5, size = 12),
        axis.title.x = element_blank(),
        axis.title.y = element_text( size = 12,face = 'bold' ),
        strip.text = element_text(size = 12,face='bold',lineheight=1),
        legend.position = "bottom",aspect.ratio = 0.8)+
  scale_fill_manual(values = colors_all)+
  guides(fill = guide_legend(override.aes = list(size=2.8),nrow=3,title=""))

myfile <- paste('plots/Figure_2D.pdf',sep ='')
pdf(myfile)
print(relative_abundance_all)
dev.off()

```


### Perform PCA using the annotated populations
In Supp_Figure_8 of the paper we show the PCA analysis at the patient level. 
Here for security and privacy reasons, we use anonymised the data and thus it is not possible to provide the code that generates the original figure shown in the paper. 
Please contact the authors to request the original data if needed. 
```{r perform PCA, echo=FALSE, cache=TRUE,eval=FALSE}



```

### Load pre-computed UMAP and visualise the populations
The UMAP found in the paper is generated using randomly subsampled cells from all patients in the cohort. However here we use anonymised data and thus it is impossible to reproduce the subsampling prodecure. 
Instead we provide pre-computed umap results without revealing personal identifiers of patients.

```{r visualiose UMAP results, echo=FALSE, cache=TRUE,eval=TRUE}

#load the pre-computed data
myfile <- paste('/Home/siv32/dkl081/Desktop/RAMMS/code_repo/anonymised_subsampled_data.RDa',sep ='')
load(myfile)

myfile <- paste('/Home/siv32/dkl081/Desktop/RAMMS/code_repo/umap_coordinates.RDa',sep ='')
load(myfile)

#markers to use for UMAP inspection
col.names <- c("CD16","CD4","CD14","CD19","CD3","CD235ab","CD11c","CD33",
               "CD133","CD123","CD162","CD185","CD45RA","CD278","CD194","CD161",
               "CD184","CD27","CD44","CD127","CD10","CD73","HLADR","CD146",
               "CD117","CD8a","CD34","CD105","CD49d","CD20","CD25","CD66b",
               "CD49f","CD45RO","CD90","CD45","CD195","CD38","CD196","CD135","CD56")

#generate the data for umap
data_umap_clust <- data_sub[ , col.names]
data_umap_clust$UMAP1 <- umap$UMAP1
data_umap_clust$UMAP2 <- umap$UMAP2
data_umap_clust$TimePoint <- data_sub$Sample
data_umap_clust$labels <- data_sub$CellType

data_umap_clust$labels <- factor(data_umap_clust$labels,levels = all_cellType_order )

umap_clust <- ggplot(data_umap_clust, aes(x = UMAP1, y = UMAP2, color = labels)) +
  geom_point(size = .1) +
  coord_fixed(ratio = 1) +
  theme_bw()+
  scale_color_manual(values = colors_all) +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        legend.position = "bottom",
        legend.background = element_rect())+
  guides(colour = guide_legend(override.aes = list(size=3), title=""))


myfile <- paste('plots/Figure_2B.pdf',sep ='')
pdf(myfile)
print(umap_clust)
dev.off()

data_umap_clust$TimePoint <- factor(data_umap_clust$TimePoint,levels = TimePointLevels)
umap_time <- ggplot(data_umap_clust, aes(x = UMAP1, y = UMAP2, color = TimePoint)) +
  geom_point(size = 0.05) +
  coord_fixed(ratio = 1) +
  scale_color_manual(values=color_TimePoints)+
  facet_wrap(~ TimePoint, ncol = 2) +
  theme_bw() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        legend.position = "bottom",aspect.ratio = 0.8)+
  guides(color = guide_legend(override.aes = list(size=3),nrow=1,title=""))

myfile <- paste('plots/UMAP_by_timePoints.pdf',sep ='')
pdf(myfile)
print(umap_time)
dev.off()


#plot the expression of markers overlaid to UMAP
color_grad_flow2 <- c("#331820", "#4d4f55", "#55626b", "#5a767e", "#628b8a", "#709b91", "#82aa96", "#98b89a", "#b0c6a2", "#c9d3ab", "#e4e0b6", "#feedc3")

image_list <- list()
myC <- 1
for(idx in col.names){
  
  tmp <- data_sub[ ,idx]
  tmp <- as.data.frame(tmp)
  data_dimred_samples <- cbind(tmp, umap)
  a <- data_sub$Sample
  #a <- as.vector(a$Sample)
  data_dimred_samples$Sample <- factor(a,levels = TimePointLevels)
  colnames(data_dimred_samples)[1] <- 'marker'
  o <- ggplot(data_dimred_samples, aes(x = UMAP1, y = UMAP2, color = marker)) +
    geom_point(size = 0.05) +
    coord_fixed(ratio = 1)+
    ggtitle(idx)+
    scale_colour_gradientn(colours = color_grad_flow2, limits = c(0,1.5),name=idx) +
    facet_wrap(~ Sample, ncol = 3) +
    theme_bw() +
    theme(axis.title.x=element_blank(),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank(),
          axis.title.y=element_blank(),
          axis.text.y=element_blank(),
          axis.ticks.y=element_blank(),
          legend.position = "bottom",aspect.ratio = 0.8)
  rm(tmp)
  image_list[[myC]] <- o
  myC <- myC + 1
}

myfile <- paste('plots//UMAP_overlaid_markers.pdf',sep ='')
pdf(myfile,onefile = T)
print(image_list)
dev.off()

```

### Generate and visualise heatmap with mean expression of the annotated populations
```{r generate and visualise heatmap, echo=FALSE, cache=TRUE,eval=TRUE}

#here we pass the cell-type identify markers used for annotation
clustering_cols <- c("CD3","CD4","CD8a","CD45RA","CD45RO",
                     "CD20","CD19","CD27",'CD117',
                     "CD56",
                     "CD127","CD34","CD38",
                     "CD16","CD14","CD11c","CD123","CD33")

# make heatmap of the annotated populations
heat_mat <- matrix(NA, nrow = length(all_cellType_order), ncol = length(clustering_cols))
myC <- 1
for(i in all_cellType_order) {
  temp_mat <- allCells[allCells$CellType == i, clustering_cols]
  heat_mat[myC,] <- as.matrix(apply(temp_mat, 2, function (x) mean(x, na.rm = T)))
  myC <- myC + 1
}

rownames(heat_mat) <- all_cellType_order
colnames(heat_mat) <- clustering_cols

# plot heatmap of expression
breaks <- seq(0, 1, by = 0.05)
white.black <- colorRampPalette(c("white", "black"))(n = 20)

#make a heatmap of all states
my_row_annot <- data.frame(CellType=rownames(heat_mat))  
rownames(my_row_annot) <- rownames(heat_mat)
my_colour <- list(
  CellType=c(CD4_Memory=colors_all[1],
             CD4_Naive=colors_all[2],
             CD8_Memory=colors_all[3],
             CD8_Naive=colors_all[4],
             T_DN=colors_all[5],
             T_other=colors_all[6],
             B_Memory=colors_all[7],
             B_Naive=colors_all[8],
             B_other=colors_all[9],
             NK_CD16_pos=colors_all[10],
             NK_CD16_neg=colors_all[11],
             NK_other=colors_all[12],
             Monocytes_classical=colors_all[13],
             Monocytes_nonClassical=colors_all[14],
             DC=colors_all[15],
             DC_plasmocytoid=colors_all[16],
             Myelocytes_other=colors_all[17]
  )
)


breaksList <- seq(0, 1, by = 0.1)
myfile <- paste('plots/Figure_1C.pdf',sep ='')
pheatmap::pheatmap(heat_mat,filename =myfile,
                   display_numbers=F,
                   fontsize_number = 6,
                   cluster_rows = F,
                   cluster_cols = F,
                   #breaks = breaksList,
                   annotation_row = my_row_annot,
                   annotation_colors = my_colour,
                   color = brewer.pal(9,'Greys'),
                   gaps_row = c(6,9,12,14),
                   gaps_col = c(5,9,10,13,15))

######################### make ridges plot with subsampled cells and fewer markers ######################### 

clustering_cols <- c("CD3","CD4","CD8a",
                     "CD20",
                     "CD56",
                     "CD16","CD14","CD123",'CD11c')

Ncell <- 1000000
a <- sample(nrow(allCells))
asel <- a[1:Ncell]

tmpLabels <- allCells[asel,'CellType']
data <- allCells[asel,clustering_cols]
data <- as.matrix(data)
rng <- colQuantiles(data, probs = c(0.01, 0.99))
data01 <- t((t(data) - rng[, 1]) / (rng[, 2] - rng[, 1]))
data01[data01 < 0] <- 0
data01[data01 > 1] <- 1
test <- data.frame(data01)
test$Class <- tmpLabels
test <- melt(test,id.vars = c('Class'))
allCells_annotated_ridgesPlot <- ggplot(test, aes(x = value, y = Class, fill = variable)) +
  geom_density_ridges(scale=1)+
  facet_wrap(~ variable, scales = "free", nrow = 4)+
  theme_ridges() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle=45, hjust=1,size = 7),
        axis.text.y = element_text(hjust=1,size = 7),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),aspect.ratio = 1.58)+
  scale_x_continuous(breaks=c(0,0.2,0.4,0.6,0.8,1))


myfile <- paste('plots/RidgesPlot_allCells.pdf',sep ='')
pdf(myfile)
print(allCells_annotated_ridgesPlot)
dev.off()

```

### Compute log2 fold changes between samples before and after aHSCT
```{r generate plots with fold changes v1,echo=FALSE}

cellFreq <- allCells %>% 
  group_by(Sample,CellType) %>% 
  summarise(samples = n())%>% 
  mutate(freq = samples/sum(samples),total=sum(samples))

group0 <- 'HC'
groupA <- 'BL'
groupD <- '100d'
groupE <- '6m'
groupF <- '12m'
groupG <- '24m'

#we compute the log fold change in abundance using as baseline the pre-aHSCT sample
summaryFC <- data.frame()
summaryFC_dable <- data_frame()
for(myCellType in all_cellType_order){
  
  dt <- cellFreq[cellFreq$CellType==myCellType,]
  
  a1 <- dt[dt$Sample==groupA,]
  a2 <- dt[dt$Sample==groupD,]
  aAD <- data.frame(Comparison='100d/BL',
                  CellType=myCellType,
                  logFC = log2(a2[,'freq']/a1[,'freq']))
                  #FC=(a2[,'freq']-a1[,'freq'])/a1[,'freq'])
  summaryFC <- rbind(summaryFC,aAD)
  
  a3 <- dt[dt$Sample==groupE,]
  aAE <- data.frame(Comparison='6m/BL',
                  CellType=myCellType,
                  logFC = log2(a3[,'freq']/a1[,'freq']))
                  #FC=(a2[,'freq']-a1[,'freq'])/a1[,'freq'])
  summaryFC <- rbind(summaryFC,aAE)
  
  a4 <- dt[dt$Sample==groupF,]
  aAF <- data.frame(Comparison='12m/BL',
                  CellType=myCellType,
                  logFC = log2(a4[,'freq']/a1[,'freq']))
                  #FC=(a2[,'freq']-a1[,'freq'])/a1[,'freq'])
  summaryFC <- rbind(summaryFC,aAF)
  
  a5 <- dt[dt$Sample==groupG,]
  aAG <- data.frame(Comparison='24m/BL',
                  CellType=myCellType,
                  logFC = log2(a5[,'freq']/a1[,'freq']))
                  #FC=(a2[,'freq']-a1[,'freq'])/a1[,'freq'])
  summaryFC <- rbind(summaryFC,aAG)
  
  #generate a table representation
  tmp <- data.frame(
                    `V5/V4` = aAD$freq,
                    `V6/V4` = aAE$freq,
                    `V7/V4` = aAF$freq,
                    `V9/V4` = aAG$freq,
                     CellType=myCellType)
  summaryFC_dable <- rbind(summaryFC_dable,tmp)
  
}

 summaryFC$CellType <- factor(summaryFC$CellType,levels = all_cellType_order)
 summaryFC$Comparison <- factor(summaryFC$Comparison , levels = c('100d/BL',
                                                                  '6m/BL',
                                                                   '12m/BL',
                                                                  '24m/BL'
                                                                  ))  
 

quantile(summaryFC$freq,probs = seq(0,1,0.05))
c3 <- quantile(summaryFC$freq,probs = seq(0,1,0.05))[3]
#threshold for the higher values
c4 <- quantile(summaryFC$freq,probs = seq(0,1,0.05))[19]
 
o <- ggplot(summaryFC,aes(x=reorder(CellType,-freq),y=freq,fill=CellType,color=CellType))+
            geom_bar(stat='identity',width = 0.58,alpha=1,size=0.2,color='black')+
            geom_hline(yintercept=c3, linetype="dashed",color = "red", size=0.2)+
            geom_hline(yintercept=c4, linetype="dashed",color = "red", size=0.2)+
            theme_bw()+
            facet_wrap(~Comparison,nrow = 1)+
            ylim(-6.5,6.5)+
            ylab('Relative abundance log2 fold change')+
             scale_fill_manual(values=colors_all)+
              theme(legend.position = "none",
                axis.title.y = element_blank(),
                axis.text.y = element_text(size = 10),
                axis.text.x = element_text(angle = 0, vjust = 0.2, hjust = 0.94, size = 10),
                axis.title.x = element_text(size = 10),
                aspect.ratio = 1.8,
                strip.text = element_text(size = 10,face='bold',lineheight=1))+coord_flip()

myfile <- paste('plots/Figure_3B.pdf',sep ='')
#png(myfile,width = 800,height = 800)
pdf(myfile)
print(o)
dev.off() 

o <- ggplot(summaryFC,aes(x=CellType,y=freq,fill=CellType,color=CellType))+
            geom_bar(stat='identity',width = 0.58,alpha=1,size=0.2,color='black')+
            geom_hline(yintercept=-2, linetype="dashed",color = "red", size=0.2)+
            geom_hline(yintercept=2, linetype="dashed",color = "red", size=0.2)+
            theme_bw()+
            facet_wrap(~Comparison,nrow = 1)+
            ylim(-6.5,6.5)+
            ylab('Relative abundance log2 fold change')+
             scale_fill_manual(values=colors_all)+
              theme(legend.position = "none",
                axis.title.y = element_blank(),
                axis.text.y = element_text(size = 10),
                axis.text.x = element_text(angle = 0, vjust = 0.2, hjust = 0.94, size = 10),
                axis.title.x = element_text(size = 10),
                aspect.ratio = 1.8,
                strip.text = element_text(size = 10,face='bold',lineheight=1))+coord_flip()


myfile <- paste('plots/Figure_3B_nonOrder.pdf',sep ='')
#png(myfile,width = 800,height = 800)
pdf(myfile)
print(o)
dev.off() 

```

### Generate summary statistics about cellular abundances per individual in the cohort
Here we use anonymised data and thus it is impossible to reproduce the intermediate results presented in the original Figure 3A. Instead we provide anonymised pre-computed data to generate the same plot. 
```{r generate summary statistics per patient,cache=TRUE,echo=FALSE,warning=FALSE}

load('/Home/siv32/dkl081/Desktop/RAMMS/code_repo/summaryStat.RDa')

#visualise the lines
summaryStat$TimePoint <- factor(summaryStat$TimePoint,levels =c(group0,groupA,groupD,groupE,groupF,groupG) )
summaryStat$CellType <- factor(summaryStat$CellType,levels = myCellTypes)

o1 <- ggplot(summaryStat, aes(TimePoint, Avg,fill=CellType,color=CellType)) +
    geom_line(aes(group = Flag),color='black',alpha=0.6) +
    geom_errorbar( aes(ymin = Avg-StdE, ymax = Avg+StdE),width = 0.15) +
    geom_point(size = 1.2)+
    facet_wrap(~CellType,nrow=3,scales = 'free_y')+
    theme_clean() +
    ylab('Mean abundance with SE')+
    scale_fill_manual(values = colors_all)+
    scale_color_manual(values = colors_all)+
    theme(axis.text.y = element_text( size = 8 ),
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.95, size = 8),
          axis.title.x = element_blank(),
          axis.title.y = element_text( size = 8 ),
          strip.text = element_text(size = 8,face='bold',lineheight=1),
          legend.position = "none",aspect.ratio = 0.8)

myfile <- paste('plots/Figure_3A.pdf',sep ='')
#png(myfile,width = 800,height = 800)
pdf(myfile)
print(o1)
dev.off() 


```

### Differential cell type abundance using diffcyt
We used the paired matrix design, which requires to use data with patient information available. However,our publicly available data are anonymised and thus it is impossible to reproduce the intermediate results presented in the volcano plot (Figure 3C). Instead we provide aggregated pre-computed data to generate the same plot. 

```{r volcano plot,cache=TRUE,echo=FALSE,eval=FALSE}
 
load('/Home/siv32/dkl081/Desktop/RAMMS/code_repo/volcanoPlotData.RDa')

o <- ggplot(data=volcanoPlot_data, aes(x=FC, y=-log10(AdjPvalue),fill=CellType,color=CellType)) +
  geom_point(size=2)+
  #geom_label_repel(aes(label=CellType),color='black',size=2,segment.size=0.1,box.padding=0.1,alpha=0.818,
  #                 max.overlaps=30)+
  geom_text(aes(label=CellType),color='black',size=1.8,check_overlap = TRUE,hjust = 0.1, nudge_x = 0.05)+
  theme_bw()+
  xlim(-6.5,6.5)+
  facet_wrap(~Comparison,scales='free_y',nrow=1)+  
  #scale_fill_manual(values = c('darkred','darkmagenta','cyan3'))+
  ylab('-log10(Adjusted p value)')+
  xlab('Relative abundance log2 fold change')+
  geom_vline(xintercept=c(c3, c4), col="red",size=0.2,linetype="dashed") +
  geom_hline(yintercept=-log10(0.05), col="red",size=0.2,linetype="dashed")+
  scale_fill_manual(values=colors_all)+
  scale_color_manual(values=colors_all)+
  theme(axis.text.y = element_text( size = 12 ),
        axis.text.x = element_text( size = 12),
        axis.title.y = element_text( size = 12 ),
        axis.title.x = element_text( size = 12 ),
        legend.position = "none", legend.text=element_text(size=12),
        panel.border = element_rect(colour = "black", fill = NA),
        legend.title = element_blank(),aspect.ratio = 1)+
  guides(fill=guide_legend(nrow=2,override.aes = list(size=2.8)))

myfile <- paste('plots/Figure_3C.pdf',sep ='')
#png(myfile,width = 800,height = 800)
pdf(myfile)
print(o)
dev.off() 
```

